### 2016-02-01 paper.tex
Initial ShareLaTeX Import

\documentclass[12pt]{article}

% sorting=none will order citations in order of appearance
% use sorting=nyt to sort by name, year, title
\usepackage[numbers, backend=biber,autolang=hyphen, bibencoding=inputenc, strict, isbn=false, uniquename=false, sorting=none, sortcites=true, style=numeric, autocite=superscript, url=false]{biblatex} % biblatex setup
\DeclareLanguageMapping{american}{american-apa}
\appto{\bibfont}{\footnotesize} % for 10pt bibliography
%\usepackage{xunicode}
\usepackage[american]{babel}
\usepackage[babel]{csquotes}
\usepackage[utf8]{inputenc}
\usepackage{float}
\usepackage{graphicx}
\usepackage[margin=1in]{geometry}
\usepackage[dvipsnames,table]{xcolor}
\usepackage{tikz}
\usetikzlibrary{positioning,arrows,shapes,fit}
\usepackage{standalone}
\usepackage{mdframed}
\usepackage{capt-of}
\usepackage{changepage}
\usepackage{titlesec}
\usepackage[parfill]{parskip}

\titleformat*{\section}{\normalsize\bfseries} % Makes section titles 12 pt font
\titleformat{\section}{\normalsize\bfseries}{}{}{}[]
\titleformat{\subsection}{\bfseries}{}{}{}[]

\let\textcite=\autocite
\addbibresource{../../computing.bib}

\title{\large \textbf{Leveraging Historical Ties between Cognitive Science and
  Computer Science to Guide Innovative Use of Computers in Education}}

% Author info isn't included for the Annual Conference but some regional conferences might request it.
\author{}
%\author{\normalsize Author Name\\
%\normalsize email@example.com\\
%\normalsize Name of Your Department\\\
%\normalsize Your Institution Name}
\date{}

\makeatletter % This gets the margins for the title set.
\patchcmd{\@maketitle}{\begin{center}}{\begin{adjustwidth}{0.5in}{0.5in}\begin{center}}{}{}
\patchcmd{\@maketitle}{\end{center}}{\end{center}\end{adjustwidth}}{}{}
\makeatother

\begin{document}
\raggedright
\maketitle
\thispagestyle{empty}
\pagestyle{empty}

% conflict between teaching programming syntax, teaching ``how to program'', teaching high-level concepts like abstractions and how to ``do'' problem solving.
\section{Coding for Everyone}
In the past few years, there has been increasing interest in a
movement to get more people, regardless of background, to learn to
program, in fact President Obama recently made a statement calling on
all children to have the opportunity to learn about computer
science\autocite{whitehouse_computer_2016}. Sites such as code.org
promote CS education opportunities for all citing statistics about
STEM jobs and arguing that in the 21st century knowledge about
computer science is foundational\autocite{code.org_every_2016}. Many
institutions of higher education either have programming requirements
for non-computer-science majors, or have been expanding programming to
non-majors\autocite{rich_cs1_2004,forte_motivation_2005,guzdial_design_2005}. At
the authors of this paper's institution, all first year general
engineering students are required to complete two 6-week rotations of
an introductory programming curriculum as part of their first-year
engineering course.

Juxtaposed with the observation that technology is changing rapidly,
and the half-life of technical skills is decreasing (cite half-life
article) it would seem teaching programming is a questionable use of
class time: the language and paradigm students are taught will likely
not be the same they are asked to use in their first job
post-graduation. However, if we acknowledge that learning to program
is more than simply learning to write code, and in particular if we
draw on work from the early era of computing which closely linked
programming to problem solving and cognitive skills, we can argue that
regardless of language or paradigm, learning to program may actually
be a way to learn to problem solve, reason about and organize
knowledge, and conceptualize processes.

While there are some recent studies which investigate programming
education as a means of teaching problem-solving, there is also a
large body of research that primarily focuses on teaching students the
details of how to code. We argue that in most cases these details are
less important than centering programming education around
problem-solving skills and if educators hope to achieve this goal then
all aspects of introductory programming courses, from the course
objectives, to the curriculum used, to the assessments administered,
must be chosen intentionally with this goal in mind.

%We posit that as long as researchers focus on only one aspect of a
%programming course, whether it be the language to use, the paradigm to
%teach, the assessment method to use, etc., there will continue to be a
%lack of con consensus of what approach is best, just as
%\textcite{pears_survey_2007-1} observed.

\section{Historic Ties Between Cognitive Science and Computer Science}
At the dawn of the computing revolution, visionaries predicted that
computers had the potential to “augment the human intellect”
\autocite{engelbart_augmenting_1962}, a much more expansive view than
simply becoming tools to automate tedious calculations. What was meant
by this was that computing allows for new ways of dealing with
complexity, provides new metaphors for thinking about
systems\autocite{kay_history_1996}. Early work demonstrated that with
the aid of computers, children could engage with tasks and concepts
that had previously been considered “advanced”, and through
interacting with computers, children could become more aware of their
own cognitive processes \autocite{papert_mindstorms_1980}. However,
\textcite{east_patternbased_1996} note that as of 1996 little work had
been done to inform instruction from knowledge of learning. The rise
in popularity of object-oriented programming seems to dominate the
conversation for the next few years, though
\textcite{robins_learning_2003} observes there is a lack of evidence
supporting any claim that OOP makes modeling problems any easier.

%“Computing enters as a new representation system with new and useful
%metaphors for dealing with complexity, especially of systems”
%\autocite{kay_history_1996}

% “I wasn't smarter but I had a much better internal thinking tool to
% amplify my abilities.” \autocite[][p.526]{kay_history_1996} about his
% ability to quickly solve a problem using LISP.

% “the 'hacker phenomenon'” \autocite[][p.544-5]{kay_history_1996}

% Difficulty novices experience have “little to do with features of the
% programming languages used, and everything to do with the difficulties
% novices have thinking the secial way that good programmers think”
% \autocite[][p.546]{kay_history_1996}

% “However, a program is not just for the computer: A program that
% follows the rules of discourse better enables a human reader to
% reconstruct the explanation that the program writer was following in
% initially developing the mechanism.”
% \autocite[][p.857]{soloway_learning_1986}

\section{Methods}
Assuming that there would be more comprehensive research of
introductory programming in the computer science community we began
with targeted searchs of the ACM digital library for `introductory
programming'. Once we found key papers that seemed to indicate a call
to action
\autocite[e.g.][]{mccracken_multinational_2001,kolling_problem_1999},
or a survey of practices
\autocite[e.g.][]{pears_survey_2007-1,robins_learning_2003} we used
the `cited by' list to find additional relevant papers.

%% Despite hope in the 1990s that computing and programming had potential
%% for bringing more undrerepresented groups into STEM fields, it has
%% become one of the *least* diverse fields.

% \textcite{turkle_epistemological_1990} were optimistic that paradigm
% shift to object oriented programming that began in the 1990s would
% result in more epistemological diversity since at its core, OOP as a
% design philosophy is compatible with what they call a “bricolage”
% strategy towards design (e. g. the concept of message passing between
% agents invokes association with a dialog or negotiation which is how
% people following a bricolage style see themselves interacting with the
% machine). Unfortunately, based on more current literature regarding
% how OOP is taught the elements of OOP that most align with a bricolage
% style are left out while concepts of OOP that align with the
% traditional “planner” style: encapsulation, inheritance are
% emphasized.

\section{Current state of research}
The bulk of literature about teaching introductory programming (CS1)
revolves around what paradigms are “best” (e.g. procedural vs. object
oriented), which language is “best”, often times with regard to a
particular paradigm (e.g. OOP with Python vs. Java), and best teaching
practices (e.g. use of “pair programming” environments). However, the
\emph{reasons} for picking one paradigm over another tend to be about
industry demand, rather than intentionally choosing a paradigm to
support the learning objectives of the course.

\section{Overall direction}
A \citeyear{robins_learning_2003} study by
\citeauthor{robins_learning_2003} identifies four trends in research
into programming education: 1) distinguishing between novice and
expert programmers with emphasis on deficiencies of novices, 2) the
distinction between knowledge and strategies, 3) the distinction
between program comprehension and generation, and 4) a comparison of
OOP to procedural styles of programming\autocite{robins_learning_2003}.

The \textcite{mccracken_multinational_2001} study, which concluded that
regardless of nation or institution, most students do not know how to
program by the end of their introductory programming courses, sparked
much conversation and push to reform introductory programming in the
next decade. Ten years after the
\citeauthor{mccracken_multinational_2001} study,
\textcite{lister_ten_2011} reflected on the response and lemented that
while much of the criticism that the problems used to test student
knowledge were too difficult, no one had repeated the study with
problems they deemed to be more reasonable. Thus while the
\citeauthor{mccracken_multinational_2001} study validated the
experiences of many introductory programming instructors it did not
result in critical change to address the problems. Rather, what we
have seen since the 1990s is been an increasing interest in where
object oriented programming should be placed within the curriculum
with debates between objects-first vs procedural-fist but little
question of why object oriented design should be included at all other
than it continues to be a popular paradigm in industry.

Much fewer, but still some, authors have drawn on cognitive sciences
to determine how indroductory programming courses should be
taught. \textcite{muller_pattern_2005} propose a pattern oriented
approach that places emphasis on recognizing and learning to work with
algorithmic patterns over syntax and design paradigms.

There remains a desire that introductory programming courses should
teach problem-solving skills, but few papers present evidence that the
implementation of this goal is informed by cognitive science.

\citeauthor{pears_survey_2007-1} conclude their
\citeyear{pears_survey_2007-1} literature survey with the observation
that despite the volume of work exploring the implementation of
introductory programming courses “there is little systematic evidence
to support any particular approach.”\autocite{pears_survey_2007-1} This
could be due to two reasons:

\begin{enumerate}
\item There has been a lack of consensus on what the goals of an
  introductory course should be: the ability to write a complete,
  working program, the ability to answer conceptual questions about
  computing and computer science, or the ability to analyze problems
  and design solutions (problem-solving). The “appropriate” approach
  to teaching would depend on the goal.

\item Even if goals were clearly stated it is likely many different
  approaches could be used successfully if chosen intentionally to
  align with the goal and measurements of success (assessment). In
  many of the studies the metrics used to determine success suffer
  from validity issues, e.g. the validity of final grades on a final
  assessment depend tremendously on the design of that assessment and
  there is little evidence to suggest course assessments are
  intentionally designed with learning goals in mind.
\end{enumerate}

A large number of papers surveyed contain discussions of course
objectives, pedagogical tools used, and assessments given, but rarely
are these three components discussed together.

\subsection{Components of a learning
system}\label{sec:components-of-a-learning-system}

A popular way to visualize an effective learning system is as a triad
of objectives, instruction, and
assessment\textcite{felder_designing_2003}, shown in Figure
\ref{fig:learning-system}. The design of these three components is not
sequential: each informs the other.

\begin{figure}[h!]
  \centering
  \includestandalone[width=\textwidth]{felder_framework}%
  \caption{Learning system}
  \label{fig:learning-system}
\end{figure}

\subsection{Effective learning systems}\label{sec:effective-learning-systems}

Assessment results can help inform changes to course objectives and
instructional methods\autocite{felder_designing_2003}. Similarly, as
instructional methods improve, the objectives may be modified to aim
for higher levels of understanding. Being intentional about these
interactions is crucial to an effective learning system; if the
assessment addresses lower levels of learning than the objectives
specify, ``the system will be driven by backwash from testing, not by
the curriculum.''\autocite[][p. 350]{biggs_enhancing_1996}. For
example, if an objective of a programming course is that learners will
be able to transfer high level concepts from programming to other
domains, but assessment only consists of checking that programs
function conforms to data input/output specifications,
teaching/learning activities would tend to center around those that
promote correct input/output behavior and exclude activities that
might have a better chance at promoting higher level understanding and
transfer\autocite{biggs_enhancing_1996}.

\section{Case Study}
\subsection{Object Oriented Programming}
As noted, teaching OOP is a common area of interest in the literature.

It should be strongly noted that we are not advocating for OOP as a
``best'' teaching paradigm. Rather, we want to point out that from what
has been reported it seems that it is often not taught
effectively. What educators seem to miss is that OOP defines an
entirely different paradigm for problem solving than is used in
procedural programming\autocite{kolling_problem_1999}. As should not
be surprising there are some problems for which OOP provides a good
set of tools to solve in an efficient, elegant manner, namely those
with solutions taht naturally map to real-world
objects\autocite{raymond_art_2003}. Different problems may be better
solved using a procedural, or functional
paradigm\autocite{kay_history_1996,felleisen_functional_2009,crestani_experience_2010}. Ultimately,
our goal should be to help students to analyze problem structures and
make informed decisions regarding what paradigm would be best for a
particular problem. What is not clear from the literature is whether
or not this consideration is made when designing problems for a
particular course. Indeed, in some cases there is evidence that OOP
concepts are added to a course ``on top of'' older content, without
necessarily changing the problem sets (citation needed).

Alan Kay originally conceived of object oriented programming as as one
way to think about complex systems\autocite{kay_history_1996}. The
underlying metaphore of Kay's conception of OOP was that of the
biological cell: each object is designed to accomplish a specific set
of tasks where the complexity of these tasks is encapsulated within
the object. Objects interact with one another by sending and receiving
messages and systems of these objects are built to solve complex
problems\autocite{kay_history_1996}.

[section on what an ideal alignment would look like using OOP]

\section{Language choice}
Choice of language does matter: If the learning objectives of an
introductory programming course are to learn problem solving concepts
then a language with a simpler syntax is preferable to one with more
complex syntax features\autocite{koulouri_teaching_2014}. In an
attempt to address the concern that the two primary languages, C++ and
Java, used to teach OOP in introductory courses, may be too complex
for novice students, \textcite{goldwasser_teaching_2008} chooses
Python to be the language of choice for introductory programming
claiming that its relatively simple syntax allow new students to
better engage with the OOP concepts rather than getting bogged down by
details of the language.

% \section{What are “the basics”}
% Several papers discuss the relative merits of a more “modern” paradigm
% such as OOP and procedural paradigms and what they refer to as “basic
% concepts” in programming. The “basics” tend to be conditional
% structures, looping structures, and arrays. However it is never noted
% that these “concepts” are only “basic” in the context of a particular
% paradigm, in particular procedural. For example, if we decided it was
% best to teach a functional paradigm in CS1 we would be hard pressed to
% defend conditional and looping structures as “basic”, rather we would
% consider recursion, evaluation, and higher-order-functions as “basic”
% concepts.

% \section{The relative dearth of functional-first approaches}
% Given the apparent focus on formalist design practices, what
% \textcite{turkle_epistemological_1990} refer to as “the planner”
% approach, it is interesting to note the relative lack of interest in
% using functional-first strategies. The functional programming paradigm
% is highly compatible with formal and abstract design methods and is
% closely associated with mathematical problem solving
% \autocite{backus_can_1978}. The relative lower interest in functional
% programming suggests that programming paradigms are chosen more for
% practical market concerns (e.g. OOP is popular in industry, adoption
% of functional styles have been slow to adoption largely due to early
% problems with speed), rather than any conscious link between design
% strategies, objectives and paradigm.

% \section{Epistomological Pluralism}
% \textcite{vilner_fundamental_2007} argues that providing “different
% ways of approaching and solving problems” should be a goal of
% introductory programming courses.

\section{Discussion}

\section{Future Work}

There continues to be a disconnect between learning sciences and how
programming paradigms and languages are selected for introductory
courses. While there is quite a bit of exploration around connections
between abstract thinking, threshold concepts, etc. and ``success'',
“success” is not consistently defined, sometimes meaning the ability
to write programs (e.g. with specifications given), sometimes focusing
more on the ability to solve problems.

\subsection{Continued Ties to Learning Sciences}

The thread of learning science influence split into two branches, one
further exploring the pattern-based approach and one exploring
threshold concepts. While developing problem-solving skills remains a
goal of CS1 only the pattern-based approach thread seems to focus on
this aspect while the researchers exploring threshold concepts are
more concerned with overall success in computer science without
specific mention of problem-solving skills.

\subsubsection{Problem Solving}

Several studies pick up the pattern-based focus of programming to
encourage problem-solving skill development from the
1990s\autocite{muller_almost_2004,muller_pattern_2005,muller_patternoriented_2007,caspersen_instructional_2007-1}.

In one of these studies, the authors draw from three concepts of
cognitive science: cognitive load theory, cognitive apprenticeship,
and the theory of worked examples, to design an introductory
programming course\autocite{caspersen_instructional_2007-1}. It is
interesting to note that they select these theories based on a human
cognition model that resulted from computing: separate working and
long term memory where learning is the process for transferming
information from working to long-term memory. The course the authors
describe teachs OOP but they never explain why the chose OOP as the
paradigm, it is assumed that this is just a given feature of an
introductory programming course.

\subsubsection{Threshold Concepts}

Some studies explore potential threshold
topics\autocite{meyer_threshold_2005} for computer science and how
they might be useful for informing course
design\autocite{mccartney_liminal_2009,boustedt_threshold_2007}. \citeauthor{boustedt_threshold_2007}
identify object oriented programming and pointers as candidate
threshold topics based on interviews with students graduating with
computer science degrees. The two concepts and so these concepts are
in the context of computer science in general rather than introductory
programming, thus there is not a clear link between these concepts and
the goal of problem solving. In addition the authors note that OOP in
particular is quite broad and suggest further work to narrow the focus
of that threshold topic to one ore more specific aspects of OOP, such
as how objects communicate between one another.

\subsubsection{Objects-first vs. X}
Much of the literature focuses on the debate between introducing
object oriented programming in CS1 or whether introductory courses
should teach a procedural style. Several researchers argue that an
object-first approach emphasices coding constructs over other goals
such as problem-solving and design
\autocite{moritz_objectsfirst_2005,muller_guidelines_2005,reges_back_2006}. \textcite{moritz_objectsfirst_2005}
propose a “design-first” approach that gets students to analyze
problems and design solutions before writing any computer code.

\autocite{vilner_fundamental_2007} found no significant difference in
student performance on questions relating to recursion and efficiency
between groups that were given an object-first curriculum vs. groups
given a procedural-first curriculum. They did observe a significant
difference between the two groups when asked to design an abstract
data type: those who were in the objects-first group performed better
than those in the procedural-first group. However, it is notable that
the students were only tested on programming knowledge, not their
ability to analyze problems and design solutions.

\textcite{reges_back_2006} takes a ``back to basics'' approach eschewing
the popular OOP and reverts his CS1 course back to a procedural style
which allows him to put the emphasis back on problem solving.

\vspace{4\baselineskip}\vspace{-\parskip} % Creaters proper 4 blank line spacing.

\printbibliography

\end{document}

### 2016-02-01 felder_framework.tex
initial commit

\documentclass[tikz]{standalone}

\usepackage{tikz}
\usetikzlibrary{positioning,arrows,shapes}

\begin{document}
  \begin{tikzpicture}[node distance = 1em, auto,
    dimension/.style={draw, rectangle, text width=4cm, align=center, minimum height=3em},
    component/.style={text width=3cm, align=center, minimum height=3em}
    ]
    \node[draw, ellipse, minimum size=30pt, text width=2cm, align=center] (students) at (0,0) {Students};
    \node[dimension, above=6em of students] (objectives) {Objectives};
    \node[dimension, below left=5em of students] (instruction) {Instruction};
    \node[dimension, below right=5em of students] (assessment) {Assessment};
    \draw[<->,>=latex,line width=1mm] ([xshift=4em]instruction.north) to (objectives);
    \draw[<->,>=latex,line width=1mm] (objectives) to ([xshift=-4em]assessment.north);
    \draw[<->,>=latex,line width=1mm] (instruction) to (assessment);

    \node[component, above=3em of objectives] (program_outcomes) {program\\ outcomes};
    \node[below left=of program_outcomes] (o1) {};
    \node[component, right=-5em of o1] (blooms) {Bloom's\\ taxonomy};
    \node[below right=of program_outcomes] (o2) {};
    \node[component, left=-5em of o2] (instructor_goals) {Instructor's\\ goals};

    \foreach \component in {blooms,program_outcomes,instructor_goals} {
      \draw[] (\component) to (objectives);
    };

    \node[component, above=3em of instruction] (IT) {Instructional technology};
    \node[below left=of IT] (i1) {};
    \node[component, right=-3em of i1] (lecture) {Lecture};
    \node[below right=of IT] (i2) {};
    \node[component, left=-3em of i2] (labs) {Labs};

    \node[component, below=3em of instruction] (other_instruction) {Other techniques};
    \node[above left=of other_instruction] (i3) {};
    \node[component, right=-4em of i3] (active) {Active \& cooperative learning};
    \node[above right=of other_instruction] (i4) {};
    \node[component, left=-4em of i4](PBL) {Problem-based learning};

    \foreach \component in {lecture,labs,other_instruction,active,PBL,IT} {
      \draw[] (\component) to (instruction);
    };

    \node[component, below=3em of assessment] (other_measures) {Other measures};
    \node[above left=of other_measures] (a1) {};
    \node[component, right=-4em of a1] (surveys) {Surveys};
    \node[above right=of other_measures] (a2) {};
    \node[component, left=-4em of a2] (tests) {Tests};
    \node[component, above=3em of assessment] (CATs) {Classroom assessment techniques};

    \foreach \component in {CATs,surveys,other_measures,tests} {
      \draw[] (\component) to (assessment);
    };
  \end{tikzpicture}  
\end{document}

### 2016-02-01 paper.tex
removed unused and error-generating keywords from biblatex line

@@
 % use sorting=nyt to sort by name, year, title
-\usepackage[numbers, backend=biber,autolang=hyphen, bibencoding=inputenc, strict, isbn=false, uniquename=false, sorting=none, sortcites=true, style=numeric, autocite=superscript, url=false]{biblatex} % biblatex setup
+\usepackage[backend=biber,autolang=hyphen, bibencoding=inputenc, isbn=false, uniquename=false, sorting=none, sortcites=true, style=numeric, autocite=superscript, url=false]{biblatex} % biblatex setup
 \DeclareLanguageMapping{american}{american-apa}
@@
 \let\textcite=\autocite
-\addbibresource{../../computing.bib}
+\addbibresource{computing.bib}
 

### 2016-02-01 paper.tex
tweaks and comments after chat with Jake

@@
 \usepackage{standalone}
-\usepackage{mdframed}
-\usepackage{capt-of}
+%\usepackage{mdframed}
 \usepackage{changepage}
@@
 Juxtaposed with the observation that technology is changing rapidly,
-and the half-life of technical skills is decreasing (cite half-life
-article) it would seem teaching programming is a questionable use of
-class time: the language and paradigm students are taught will likely
-not be the same they are asked to use in their first job
-post-graduation. However, if we acknowledge that learning to program
-is more than simply learning to write code, and in particular if we
-draw on work from the early era of computing which closely linked
-programming to problem solving and cognitive skills, we can argue that
-regardless of language or paradigm, learning to program may actually
-be a way to learn to problem solve, reason about and organize
-knowledge, and conceptualize processes.
+and the half-life of technical skills is
+decreasing\autocite{nae_educating_2004} it would seem teaching
+programming is a questionable use of class time: the language and
+paradigm students are taught will likely not be the same they are
+asked to use in their first job post-graduation. However, if we
+acknowledge that learning to program is more than simply learning to
+write code, and in particular if we draw on work from the early era of
+computing which closely linked programming to problem solving and
+cognitive skills, we can argue that regardless of language or
+paradigm, learning to program may actually be a way to learn to
+problem solve, reason about and organize knowledge, and conceptualize
+processes.
 
@@
 objectives, to the curriculum used, to the assessments administered,
-must be chosen intentionally with this goal in mind.
+must be chosen intentionally with this goal in mind. In this paper we
+will present a brief summary of current published work on introductory programming education, then note a potential lack that, 
+then offer a potential framework to
 
@@
 \section{Methods}
+[identify learning sciences sources]
+
 Assuming that there would be more comprehensive research of
@@
 \section{Current state of research}
+%%TODO: add citations
 The bulk of literature about teaching introductory programming (CS1)
-revolves around what paradigms are “best” (e.g. procedural vs. object
-oriented), which language is “best”, often times with regard to a
+revolves around what paradigms are ``best'' (e.g. procedural vs. object
+oriented), which language is ``best'', often times with regard to a
 particular paradigm (e.g. OOP with Python vs. Java), and best teaching
@@
 support the learning objectives of the course.
+%% TODO: reminder to connect to learning science, e.g. this behavior
+%% is not uncommong, for instance flipped classrooms is a ``hot
+%% topic'' now and often times we see people implementing a ``flipped
+%% classroom'' without necessarily considering if that pedagogical
+%% approach makes the most sense for a given course.
 
@@
 \citeauthor{mccracken_multinational_2001} study,
-\textcite{lister_ten_2011} reflected on the response and lemented that
+\textcite{lister_ten_2011} reflected on the response and lamented that
 while much of the criticism that the problems used to test student
 knowledge were too difficult, no one had repeated the study with
-problems they deemed to be more reasonable. Thus while the
-\citeauthor{mccracken_multinational_2001} study validated the
-experiences of many introductory programming instructors it did not
-result in critical change to address the problems. Rather, what we
-have seen since the 1990s is been an increasing interest in where
-object oriented programming should be placed within the curriculum
-with debates between objects-first vs procedural-fist but little
-question of why object oriented design should be included at all other
-than it continues to be a popular paradigm in industry.
+problems they deemed to be more reasonable. 
+%% TODO: connect mccracken to engineering ed about missconceptions
+%% (CHEER chapter and Streveler paper)
+
+Thus while the \citeauthor{mccracken_multinational_2001} study
+validated the experiences of many introductory programming instructors
+it did not result in critical change to address the problems. Rather,
+what we have seen since the 1990s is been an increasing interest in
+where object oriented programming should be placed within the
+curriculum with debates between objects-first vs procedural-fist but
+little question of why object oriented design should be included at
+all other than it continues to be a popular paradigm in industry.
 
@@
 introductory programming courses “there is little systematic evidence
-to support any particular approach.”\autocite{pears_survey_2007-1} This
-could be due to two reasons:
+to support any particular approach.”\autocite{pears_survey_2007-1}
+We think this could be due to two reasons:
 
@@
 
+%%TODO: link to knowledge/skills interaction literature
+%%TODO: bring in knowledge organization, focus on concepts vs. procedures
+
 [section on what an ideal alignment would look like using OOP]
@@
 cognition model that resulted from computing: separate working and
-long term memory where learning is the process for transferming
+long term memory where learning is the process for transforming
 information from working to long-term memory. The course the authors
@@
 introductory programming course.
+
+%%TODO: consider focusing on this and dropping threshold concepts and others. Potentally bring in pattern-based pedagogy
 

### 2016-02-02 paper.tex
another wave of edits and additions
still need to rework discussion, future work, and add a conclusion.

@@
 \titleformat{\subsection}{\bfseries}{}{}{}[]
+\titleformat{\subsubsection}{\bfseries}{}{}{}[]
 
@@
 \section{Coding for Everyone}
-In the past few years, there has been increasing interest in a
-movement to get more people, regardless of background, to learn to
-program, in fact President Obama recently made a statement calling on
-all children to have the opportunity to learn about computer
+In the past few years, there has been increasing interest to get more
+people, regardless of background, to learn to program. In fact
+President Obama recently made a statement calling on all children to
+have the opportunity to learn about computer
 science\autocite{whitehouse_computer_2016}. Sites such as code.org
-promote CS education opportunities for all citing statistics about
+promote CS education opportunities for all, citing statistics about
 STEM jobs and arguing that in the 21st century knowledge about
@@
 the authors of this paper's institution, all first year general
-engineering students are required to complete two 6-week rotations of
-an introductory programming curriculum as part of their first-year
-engineering course.
+engineering students are required to complete an introductory
+programming module as part of their first-year engineering course.
 
@@
 and the half-life of technical skills is
-decreasing\autocite{nae_educating_2004} it would seem teaching
+decreasing\autocite{nae_educating_2004}, it would seem that teaching
 programming is a questionable use of class time: the language and
@@
 paradigm, learning to program may actually be a way to learn to
-problem solve, reason about and organize knowledge, and conceptualize
-processes.
+problem solve, organize knowledge, and conceptualize processes.
 
@@
 less important than centering programming education around
-problem-solving skills and if educators hope to achieve this goal then
-all aspects of introductory programming courses, from the course
-objectives, to the curriculum used, to the assessments administered,
-must be chosen intentionally with this goal in mind. In this paper we
-will present a brief summary of current published work on introductory programming education, then note a potential lack that, 
-then offer a potential framework to
+problem-solving skills. If we make learning problem-solving skills a
+goal of introductory programming courses, then all aspects of the
+course, from the course objectives, to the curriculum used, to the
+assessments administered, must be chosen intentionally with this goal
+in mind.
+
+In this paper we will present a brief summary of current published
+work on introductory programming education. We will note that much of
+the current work lacks a tight integration of course objectives,
+pedagogical strategies, and assessments, and will offer a potential
+framework to help think about both the design and evaluation of
+introductory programming courses.
 
@@
 \section{Methods}
-[identify learning sciences sources]
-
 Assuming that there would be more comprehensive research of
@@
 the `cited by' list to find additional relevant papers.
+
+We needed a framework to consistently evaluate and starting from
+\citeauthor{jamieson_creating_2009}'s
+(\citeyear{jamieson_creating_2009}) to ground educational research
+activities in how people learn we identified the concept of
+\empha{constructive alignment} as a promising one. Following the
+references in \citeauthor{jamieson_creating_2009}'s report as well as
+conducting targeted literature searches for ``constructive alignment''
+we were able to find descriptions of this
+framework\autocite{jamieson_creating_2009,pellegrino_rethinking_2006,felder_designing_2003,biggs_enhancing_1996}.
 
@@
 \section{Current state of research}
-%%TODO: add citations
 The bulk of literature about teaching introductory programming (CS1)
-revolves around what paradigms are ``best'' (e.g. procedural vs. object
-oriented), which language is ``best'', often times with regard to a
-particular paradigm (e.g. OOP with Python vs. Java), and best teaching
-practices (e.g. use of “pair programming” environments). However, the
-\emph{reasons} for picking one paradigm over another tend to be about
-industry demand, rather than intentionally choosing a paradigm to
-support the learning objectives of the course.
+revolves around what paradigms are ``best'' (e.g. procedural
+vs. object oriented
+\autocite{bailie_objects_2003,moritz_objectsfirst_2005,reges_back_2006}),
+which language is ``best'', often times with regard to a particular
+paradigm (e.g. OOP with Python
+vs. Java\autocite{goldwasser_teaching_2008,koulouri_teaching_2014}),
+and best teaching practices (e.g. use of “pair programming”
+environments\autocite{williams_support_2001,mcdowell_effects_2002,thomas_code_2003,chong_social_2007,salleh_empirical_2011}). However,
+the \emph{reasons} for picking one paradigm over another tend to be
+about industry demand, rather than intentionally choosing a paradigm
+to support the learning objectives of the course. This behavior,
+making a popular practice part of a learning environment without
+critically analyzing whether or not is a good fit is by no means
+exclusive to programming education. For example, flipped
+classrooms\autocite{bishop_flipped_2013}, in which students are asked
+to watch recorded lectures at home and use class time for working on
+activities, has recently been a hot topic in higher education and
+there have been a continuous number of examples of well-intentioned
+practitioners flipping their classroom without adjusting course
+objectives, content and activities to work under the flipped-classroom
+model\autocite{desantis_students_2015}. Likewise, anyone can teach
+elements of object oriented programming in their introductory
+programming courses, but unless critical aspects of the course are
+adjusted: background reading, rethinking what the ``fundamental
+concepts'' are, the types of problems assigned, and methods of
+assessment, the incorporation of OOP will likely create confusion for
+students and frustration for instructor\autocite{kolling_problem_1999}.
+
 %% TODO: reminder to connect to learning science, e.g. this behavior
@@
 between program comprehension and generation, and 4) a comparison of
-OOP to procedural styles of programming\autocite{robins_learning_2003}.
+OOP to procedural styles of
+programming\autocite{robins_learning_2003}. They found that the
+biggest challenge for novice programmers was not understanding the
+basic concepts, but knowing how to apply them. This finding is
+consistent understanding of how we
+learn\autocite{bransford_how_2000,ambrose_chapter_2010}, and with
+studies of expert/novice behavior in other domains, such a
+statics\autocite{litzinger_cognitive_2010}. In both cases the
+suggestion is that helping students learn strategies that work is
+crucial to helping them become effective novices, i.e. devote
+instructional time to problem-solving strategies and how to apply
+them.
 
-The \textcite{mccracken_multinational_2001} study, which concluded that
+The \citeyear{mccracken_multinational_2001}
+\citeauthor{mccracken_multinational_2001}
+study\autocite{mccracken_multinational_2001}, which concluded that
 regardless of nation or institution, most students do not know how to
@@
 much conversation and push to reform introductory programming in the
-next decade. Ten years after the
-\citeauthor{mccracken_multinational_2001} study,
-\textcite{lister_ten_2011} reflected on the response and lamented that
-while much of the criticism that the problems used to test student
-knowledge were too difficult, no one had repeated the study with
-problems they deemed to be more reasonable. 
+next decade. This phenomenon is by no means exclusive to introductory
+programming: disconnect between what we think we are teaching and what
+students learn has been observed across
+engineering\autocite{streveler_learning_2008,flynn_engineering_2014}. Possible
+ways of addressing this disconnect include recognizing differences in
+student's prior knowledge, helping students see connections between
+abstract concepts, and relating difficult concepts to ones students
+already know\autocite{streveler_learning_2008,slotta_helping_2006}.
 %% TODO: connect mccracken to engineering ed about missconceptions
@@
 
-Thus while the \citeauthor{mccracken_multinational_2001} study
-validated the experiences of many introductory programming instructors
-it did not result in critical change to address the problems. Rather,
-what we have seen since the 1990s is been an increasing interest in
-where object oriented programming should be placed within the
-curriculum with debates between objects-first vs procedural-fist but
-little question of why object oriented design should be included at
-all other than it continues to be a popular paradigm in industry.
+While the \citeauthor{mccracken_multinational_2001} study validated the
+experiences of many introductory programming instructors it did not
+result in critical change to address the problems. Rather, what we
+have seen since the 1990s is been an increasing interest in where
+object oriented programming should be placed within the curriculum
+with debates between objects-first vs procedural-fist but little
+question of why object oriented design should be included at all other
+than it continues to be a popular paradigm in industry.
 
@@
 approach that places emphasis on recognizing and learning to work with
-algorithmic patterns over syntax and design paradigms.
+algorithmic patterns over syntax and design paradigms. This approach
+closely aligns with understanding of differences in expert/novice
+behavior by helping students identify patterns of solutions and
+learning how and when to apply them to particular problems.
 
 There remains a desire that introductory programming courses should
-teach problem-solving skills, but few papers present evidence that the
-implementation of this goal is informed by cognitive science.
+teach problem-solving skills, though only a few papers present
+evidence that the implementation of this goal is informed by cognitive
+science\autocite{kramarski_cognitivemetacognitive_1997,reed_incorporating_1998,muller_supporting_2008}.
 
@@
 
-A large number of papers surveyed contain discussions of course
-objectives, pedagogical tools used, and assessments given, but rarely
-are these three components discussed together.
+Finally, a \citeyear{koulouri_teaching_2014} study by
+\citeauthor{koulouri_teaching_2014} conducted a quantitative analysis
+of different approaches to teaching introductory programming by
+varying three aspects: the language used, providing formative
+feedback, and introducing explicit problem solving
+training\autocite{koulouri_teaching_2014}. They found that choosing a
+language with a simpler syntax (Python over Java), and introducing
+explicit problem-solving training improved student learning, while
+providing more formative assessment did not seem to help. They
+speculate that the reason for the latter finding could be that
+students do not spend time looking at formative feedback, or do not
+understand what is given.
+
+We wish to build upon the past reviews of the field by suggesting an
+approach to designing learning systems that may more consistently
+address some of the shortcomings observed, and bring understanding to
+the successes. A large number of papers surveyed contain discussions
+of course objectives, pedagogical tools used, and assessments given,
+but rarely are these three components discussed together. We know from
+the engineering education literature that an effective learning
+system depends on the alignment of these three
+components\autocite{jamieson_creating_2009,pellegrino_rethinking_2006}.
 
@@
 example, if an objective of a programming course is that learners will
-be able to transfer high level concepts from programming to other
-domains, but assessment only consists of checking that programs
-function conforms to data input/output specifications,
-teaching/learning activities would tend to center around those that
-promote correct input/output behavior and exclude activities that
-might have a better chance at promoting higher level understanding and
-transfer\autocite{biggs_enhancing_1996}.
+develop problem solving skills, but assessment only consist of test of
+programming concepts and syntax, teaching/learning activities would
+tend to center around those that promote understanding of these
+concepts and exclude activities that might have a better chance at
+promoting general problem solving\autocite{biggs_enhancing_1996}.
 
 \section{Case Study}
-\subsection{Object Oriented Programming}
 As noted, teaching OOP is a common area of interest in the literature.
@@
 It should be strongly noted that we are not advocating for OOP as a
-``best'' teaching paradigm. Rather, we want to point out that from what
-has been reported it seems that it is often not taught
+``best'' teaching paradigm. Rather, we want to point out that from
+what has been reported it seems that it is often not taught
 effectively. What educators seem to miss is that OOP defines an
@@
 concepts are added to a course ``on top of'' older content, without
-necessarily changing the problem sets (citation needed).
+necessarily changing the problem
+sets\autocite{kolling_problem_1999,vilner_fundamental_2007}.
 
-Alan Kay originally conceived of object oriented programming as as one
-way to think about complex systems\autocite{kay_history_1996}. The
-underlying metaphore of Kay's conception of OOP was that of the
-biological cell: each object is designed to accomplish a specific set
-of tasks where the complexity of these tasks is encapsulated within
-the object. Objects interact with one another by sending and receiving
-messages and systems of these objects are built to solve complex
+\subsection{Object Oriented Programming}
+
+Alan Kay originally conceived of object oriented programming as
+\emph{one} way to think about complex
+systems\autocite{kay_history_1996}. The underlying metaphore of Kay's
+conception of OOP was that of the biological cell: each object is
+designed to accomplish a specific set of tasks where the complexity of
+these tasks is encapsulated within the object. Objects interact with
+one another by sending and receiving messages, and systems of these
+objects are assembled to solve complex
 problems\autocite{kay_history_1996}.
@@
 
+In several reports of attempts to incorporate OOP into introductory
+programming courses, practitioners worried that time devoted to OOP
+concepts would necessarily reduce time spent on the ``basics of
+programming'', usually defined as conditional statements, loops, and
+in some cases pointers. This concern is misplaced, under an OOP
+paradigm the ``basics of programming'' are not the same as when
+working from a procedural paradigm. OOP defines a completely different
+set of tools for problem solving, and it is these tools and concepts
+that become the
+``basics''\autocite{kolling_problem_1999,kay_history_1996}.
+
+For instance, if taking a patterns-based approach to teaching
+programming one could use either a procedural paradigm or an OOP
+paradigm, but the fundamental patterns would be quite different
+depending on the paradigm\autocite{wallingford_first_1996}.
+
+Ultimately, if our goal is to teach problem solving in general, we
+would be doing a grave disservice to our students to leave with the
+impression that there is only one paradigm, whether it be OOP,
+procedural, functional, or something not yet invented. As previously
+mentioned, OOP and procedural paradigms are comprised of fundamentally
+different concepts for problem solving, and so again is the functional
+paradigm. Unsurprisingly, there is no paradigm that is the ``best''
+fit for every problem. Leaving students with the believe that only one
+paradigm exists makes it very difficult for them to learn how to
+intentionally select and apply the ``best'' paradigm to solve a
+particular problem. \citeauthor{kolling_problem_1999} argues that
+performing a paradigm shift from procedural style to OOP is difficult,
+and so if OOP is a desired outcome then this should be taught
+\emph{first}.
+
 \section{Language choice}
@@
 then a language with a simpler syntax is preferable to one with more
-complex syntax features\autocite{koulouri_teaching_2014}. In an
-attempt to address the concern that the two primary languages, C++ and
-Java, used to teach OOP in introductory courses, may be too complex
-for novice students, \textcite{goldwasser_teaching_2008} chooses
-Python to be the language of choice for introductory programming
-claiming that its relatively simple syntax allow new students to
-better engage with the OOP concepts rather than getting bogged down by
-details of the language.
+complex syntax features\autocite{koulouri_teaching_2014}. The
+explaination for this can be found in cognitive load theory (CLT),
+which posits that learners have a finite capacity for their short-term
+memory, and for learning to occur, all aspects of the task must fit
+into this finite space. While some contributors to cognitive load are
+necessary for the task at hand, others are
+not\autocite{paas_cognitive_2003}. If, for example the goal is to
+teach problem-solving techniques there will be cognitive load
+associated with thinking about different strategies one could
+use. Learning a programming language's syntax rules also adds to the
+cognitive load, but it does not aid in the process of learning
+problem-solving techniques. Thus, it is in the educators best interest
+to minimize cognitive load associated with features that do not
+directly help with the learning objectives of the course.
+
+In an attempt to address the concern that the two primary languages,
+C++ and Java, used to teach OOP in introductory courses, may be too
+complex for novice students, \textcite{goldwasser_teaching_2008}
+chooses Python to be the language of choice for introductory
+programming claiming that its relatively simple syntax allow new
+students to better engage with the OOP concepts rather than getting
+bogged down by details of the language.
+
+\section{Problem choice}
+As previously noted, OOP provides a set of tools that naturally lend
+themselves to solving certain types of problems, but not others. An
+educator teaching OOP should carefully select problems and challenges
+that make sense to be solved with an OOP paradigm. Of course,
+provoiding illustrative examples of problems that are \emph{not} best
+solved with an OOP approach would be useful provided there is also
+time for discussion of \emph{why} OOP might not be suited for certain
+problems and brief mention of alternative approaches.
 

### 2016-02-03 paper.tex
Updates from ShareLaTeX
another wave of edits
filling in some gaps, adding future work and conclusion

@@
 \usepackage[parfill]{parskip}
+\usepackage{mdframed}
 
@@
 
-\let\textcite=\autocite
+%%\let\textcite=\autocite
 \addbibresource{computing.bib}
@@
 \title{\large \textbf{Leveraging Historical Ties between Cognitive Science and
-  Computer Science to Guide Innovative Use of Computers in Education}}
+  Computer Science to Guide Programming Education}}
 
@@
 
+%% TODO: be more explicit about learning framework
+
+%% TODO: bring in formative assessment
+
 \begin{document}
@@
 % conflict between teaching programming syntax, teaching ``how to program'', teaching high-level concepts like abstractions and how to ``do'' problem solving.
-\section{Coding for Everyone}
+\section{Coding for everyone}
 In the past few years, there has been increasing interest to get more
-people, regardless of background, to learn to program. In fact
+people, regardless of background, to learn to program. In fact,
 President Obama recently made a statement calling on all children to
@@
 engineering students are required to complete an introductory
-programming module as part of their first-year engineering course.
+programming module as part of their first-year engineering course, and
+in general ``basic programming skills'' is a common desired outcome
+across engineering
+curricula\autocite{reid_classification_2014}.
 
-Juxtaposed with the observation that technology is changing rapidly,
-and the half-life of technical skills is
-decreasing\autocite{nae_educating_2004}, it would seem that teaching
-programming is a questionable use of class time: the language and
-paradigm students are taught will likely not be the same they are
+This coding-for-all paradigm juxtaposed with the observation that
+technology is changing rapidly, and the half-life of technical skills
+is decreasing\autocite{nae_educating_2004}, it would seem that
+teaching programming is a questionable use of class time: the language
+and paradigm students are taught will likely not be the same they are
 asked to use in their first job post-graduation. However, if we
@@
 problem solve, organize knowledge, and conceptualize processes.
+
+At the dawn of the computing revolution, visionaries predicted that
+computers had the potential to ``augment the human intellect''
+\autocite{engelbart_augmenting_1962}, a much more expansive view than
+simply becoming tools to automate tedious calculations. What was meant
+by this was that computing allows for new ways of dealing with
+complexity, provides new metaphors for thinking about
+systems\autocite{kay_history_1996}. Early work demonstrated that with
+the aid of computers, children could engage with tasks and concepts
+that had previously been considered ``advanced'', and through
+interacting with computers, children could become more aware of their
+own cognitive processes \autocite{papert_mindstorms_1980}. Thus, the
+fact that specific technologies may be obsolete by the time students
+graduate may be beside the point, there are lasting cognitive benefits
+learning to interact with a computer.
 
@@
 In this paper we will present a brief summary of current published
-work on introductory programming education. We will note that much of
-the current work lacks a tight integration of course objectives,
-pedagogical strategies, and assessments, and will offer a potential
-framework to help think about both the design and evaluation of
-introductory programming courses.
-
-%We posit that as long as researchers focus on only one aspect of a
-%programming course, whether it be the language to use, the paradigm to
-%teach, the assessment method to use, etc., there will continue to be a
-%lack of con consensus of what approach is best, just as
-%\textcite{pears_survey_2007-1} observed.
-
-\section{Historic Ties Between Cognitive Science and Computer Science}
-At the dawn of the computing revolution, visionaries predicted that
-computers had the potential to “augment the human intellect”
-\autocite{engelbart_augmenting_1962}, a much more expansive view than
-simply becoming tools to automate tedious calculations. What was meant
-by this was that computing allows for new ways of dealing with
-complexity, provides new metaphors for thinking about
-systems\autocite{kay_history_1996}. Early work demonstrated that with
-the aid of computers, children could engage with tasks and concepts
-that had previously been considered “advanced”, and through
-interacting with computers, children could become more aware of their
-own cognitive processes \autocite{papert_mindstorms_1980}. However,
-\textcite{east_patternbased_1996} note that as of 1996 little work had
-been done to inform instruction from knowledge of learning. The rise
-in popularity of object-oriented programming seems to dominate the
-conversation for the next few years, though
-\textcite{robins_learning_2003} observes there is a lack of evidence
-supporting any claim that OOP makes modeling problems any easier.
+work on introductory programming education. In particular we will
+explore the alignment of course objectives, pedagogical strategies,
+and assessments, and will offer a potential framework to help think
+about both the design and evaluation of introductory programming
+courses.
 
@@
 
-We needed a framework to consistently evaluate and starting from
+Because the literature search of introductory programming seemed to be
+interest in why and how programming was taught, we needed a documented
+framework to consistently evaluate and starting from
 \citeauthor{jamieson_creating_2009}'s
@@
 activities in how people learn we identified the concept of
-\empha{constructive alignment} as a promising one. Following the
+\emph{constructive alignment} as a promising one. Following the
 references in \citeauthor{jamieson_creating_2009}'s report as well as
@@
 framework\autocite{jamieson_creating_2009,pellegrino_rethinking_2006,felder_designing_2003,biggs_enhancing_1996}.
+%% TODO: mention that we chose the framework after analyzing the above
+%% citations, i.e. it wasn't a direct pull from jamieson
 
@@
 
-%% TODO: reminder to connect to learning science, e.g. this behavior
-%% is not uncommong, for instance flipped classrooms is a ``hot
-%% topic'' now and often times we see people implementing a ``flipped
-%% classroom'' without necessarily considering if that pedagogical
-%% approach makes the most sense for a given course.
-
-\section{Overall direction}
+%% TODO: more directed title
+%% TODO: possibly use subheadings to ``tweet'' the major takeaways from each section,
+%% this could then be connected at the end to bullet points of suggestions
+\section{Trends in programming education research}
 A \citeyear{robins_learning_2003} study by
 \citeauthor{robins_learning_2003} identifies four trends in research
-into programming education: 1) distinguishing between novice and
-expert programmers with emphasis on deficiencies of novices, 2) the
+of programming education: 1) distinguishing between novice and expert
+programmers with emphasis on deficiencies of novices, 2) the
 distinction between knowledge and strategies, 3) the distinction
@@
 learn\autocite{bransford_how_2000,ambrose_chapter_2010}, and with
-studies of expert/novice behavior in other domains, such a
+studies of expert/novice behavior in other domains, such as
 statics\autocite{litzinger_cognitive_2010}. In both cases the
@@
 
-The \citeyear{mccracken_multinational_2001}
-\citeauthor{mccracken_multinational_2001}
-study\autocite{mccracken_multinational_2001}, which concluded that
-regardless of nation or institution, most students do not know how to
-program by the end of their introductory programming courses, sparked
-much conversation and push to reform introductory programming in the
-next decade. This phenomenon is by no means exclusive to introductory
-programming: disconnect between what we think we are teaching and what
-students learn has been observed across
+A \citeyear{mccracken_multinational_2001} study by
+\citeauthor{mccracken_multinational_2001}\autocite{mccracken_multinational_2001},
+which concluded that regardless of nation or institution, most
+students do not know how to program by the end of their introductory
+programming courses, sparked much conversation and push to reform
+introductory programming in the next decade. This phenomenon is by no
+means exclusive to introductory programming: disconnect between what
+we think we are teaching and what students learn has been observed
+across
 engineering\autocite{streveler_learning_2008,flynn_engineering_2014}. Possible
@@
 
-Much fewer, but still some, authors have drawn on cognitive sciences
-to determine how indroductory programming courses should be
-taught. \textcite{muller_pattern_2005} propose a pattern oriented
-approach that places emphasis on recognizing and learning to work with
-algorithmic patterns over syntax and design paradigms. This approach
-closely aligns with understanding of differences in expert/novice
-behavior by helping students identify patterns of solutions and
-learning how and when to apply them to particular problems.
-
 There remains a desire that introductory programming courses should
@@
 evidence that the implementation of this goal is informed by cognitive
-science\autocite{kramarski_cognitivemetacognitive_1997,reed_incorporating_1998,muller_supporting_2008}.
+science\autocite{kramarski_cognitivemetacognitive_1997,reed_incorporating_1998,muller_supporting_2008}. One
+thread in particular has received regular attention:
+\citeauthor{east_patternbased_1996} propose a pattern-based approach
+that places emphasis on recognizing and learning to work with
+algorithmic patterns over syntax and design
+paradigms\autocite{east_patternbased_1996}. This approach closely
+aligns with understanding of differences in expert/novice behavior by
+helping students identify patterns of solutions and learning how and
+when to apply them to particular
+problems\autocite{east_patternbased_1996,muller_pattern_2005,wallingford_first_1996}.
 
@@
 
-\item Even if goals were clearly stated it is likely many different
-  approaches could be used successfully if chosen intentionally to
-  align with the goal and measurements of success (assessment). In
-  many of the studies the metrics used to determine success suffer
-  from validity issues, e.g. the validity of final grades on a final
-  assessment depend tremendously on the design of that assessment and
-  there is little evidence to suggest course assessments are
-  intentionally designed with learning goals in mind.
+\item Even if goals were clearly stated it is likely that many
+  different approaches could be used successful if chosen
+  intentionally to align with the goal and measurements of success
+  (assessment). In many of the studies it is not immediately clear if
+  the measures of success used are appropriate for the stated
+  objectives. That is to say, the assessments used might not always be
+  measuring all aspects of the desired learning objectives.
 \end{enumerate}
 
-Finally, a \citeyear{koulouri_teaching_2014} study by
+Finally, in \citeyear{koulouri_teaching_2014},
 \citeauthor{koulouri_teaching_2014} conducted a quantitative analysis
@@
 training\autocite{koulouri_teaching_2014}. They found that choosing a
-language with a simpler syntax (Python over Java), and introducing
-explicit problem-solving training improved student learning, while
-providing more formative assessment did not seem to help. They
-speculate that the reason for the latter finding could be that
-students do not spend time looking at formative feedback, or do not
-understand what is given.
+language with a simpler syntax (Python instead of Java), and
+introducing explicit problem-solving training improved student
+learning, while providing more formative assessment did not seem to
+help. They speculate that the reason for the latter finding could be
+that students do not spend time looking at formative feedback, or do
+not understand what is given.
 
 We wish to build upon the past reviews of the field by suggesting an
-approach to designing learning systems that may more consistently
+approach to designing learning systems that may help to consistently
 address some of the shortcomings observed, and bring understanding to
@@
 but rarely are these three components discussed together. We know from
-the engineering education literature that an effective learning
-system depends on the alignment of these three
+the engineering education literature that an effective learning system
+depends on the alignment of these three
 components\autocite{jamieson_creating_2009,pellegrino_rethinking_2006}.
@@
 of objectives, instruction, and
-assessment\textcite{felder_designing_2003}, shown in Figure
+assessment\autocite{felder_designing_2003}, shown in Figure
 \ref{fig:learning-system}. The design of these three components is not
@@
   \includestandalone[width=\textwidth]{felder_framework}%
-  \caption{Learning system}
+  \caption{Learning system as derived from
+    \citeauthor{felder_designing_2003}\autocite{felder_designing_2003}}
+  %%TODO: as derived from original figure
   \label{fig:learning-system}
@@
 
-\section{Case Study}
-As noted, teaching OOP is a common area of interest in the literature.
+\section{A hypothetical case study}
+As noted, teaching OOP is a common area of interest in the
+literature. While \citeauthor{robins_learning_2003} note there is a
+lack of evidence supporting any claim that OOP makes modeling problems
+any easier\autocite{robins_learning_2003}, we will incorporate it into
+a hypothetical introductory programming course to explore how its
+inclusion would affect the constructive alignment between course
+objectives, instructional methods, and assessment.
 
@@
 ``best'' teaching paradigm. Rather, we want to point out that from
-what has been reported it seems that it is often not taught
-effectively. What educators seem to miss is that OOP defines an
-entirely different paradigm for problem solving than is used in
-procedural programming\autocite{kolling_problem_1999}. As should not
-be surprising there are some problems for which OOP provides a good
-set of tools to solve in an efficient, elegant manner, namely those
-with solutions taht naturally map to real-world
+what has been reported teaching OOP effectively can be
+challenging. What many implementations seem to miss is that OOP
+defines an entirely different paradigm for problem solving than is
+used in procedural programming\autocite{kolling_problem_1999}. As
+should not be surprising there are some problems for which OOP
+provides a good set of tools to solve in an efficient, elegant manner,
+namely those with solutions that naturally map to real-world
 objects\autocite{raymond_art_2003}. Different problems may be better
@@
 paradigm\autocite{kay_history_1996,felleisen_functional_2009,crestani_experience_2010}. Ultimately,
-our goal should be to help students to analyze problem structures and
+our goal should be to help students analyze problem structures and
 make informed decisions regarding what paradigm would be best for a
@@
 
-\subsection{Object Oriented Programming}
+\subsection{Object oriented programming}
 
@@
 %%TODO: bring in knowledge organization, focus on concepts vs. procedures
-
-[section on what an ideal alignment would look like using OOP]
 
@@
 performing a paradigm shift from procedural style to OOP is difficult,
-and so if OOP is a desired outcome then this should be taught
-\emph{first}.
+and so concludes that if OOP is a desired outcome then it should be
+taught \emph{first} since the reverse shift, from OOP to procedural,
+is conceptually easier.
 
-\section{Language choice}
+\subsection{Language choice}
 Choice of language does matter: If the learning objectives of an
@@
 complex syntax features\autocite{koulouri_teaching_2014}. The
-explaination for this can be found in cognitive load theory (CLT),
+explanation for this can be found in cognitive load theory (CLT),
 which posits that learners have a finite capacity for their short-term
@@
 necessary for the task at hand, others are
-not\autocite{paas_cognitive_2003}. If, for example the goal is to
-teach problem-solving techniques there will be cognitive load
+not\autocite{paas_cognitive_2003}. If, for example, the goal is to
+teach problem-solving techniques, there will be cognitive load
 associated with thinking about different strategies one could
-use. Learning a programming language's syntax rules also adds to the
+use. Learning the syntax rules of a programming language adds to the
 cognitive load, but it does not aid in the process of learning
-problem-solving techniques. Thus, it is in the educators best interest
-to minimize cognitive load associated with features that do not
-directly help with the learning objectives of the course.
+problem-solving techniques. Thus, it is in the educator's best
+interest to minimize cognitive load associated with features that do
+not directly help with the learning objectives of the course and
+select a teaching language with easy-to-learn syntax.
 
@@
 C++ and Java, used to teach OOP in introductory courses, may be too
-complex for novice students, \textcite{goldwasser_teaching_2008}
-chooses Python to be the language of choice for introductory
-programming claiming that its relatively simple syntax allow new
-students to better engage with the OOP concepts rather than getting
-bogged down by details of the language.
+complex for novice students, \citeauthor{goldwasser_teaching_2008}
+select Python as a good choice for introductory programming claiming
+that its relatively simple syntax allow new students to better engage
+with the OOP concepts rather than getting bogged down by details of
+the language\autocite{goldwasser_teaching_2008}.
 
-\section{Problem choice}
+\subsection{Problem Solving}
+Developing problem solving was widely regarded as a common objective
+for introductory programming courses. Achieving this objective is more
+likely if the course begins with a focus on problem solving
+strategies, before any programming is
+introduced\autocite{koulouri_teaching_2014}. Separating
+problem-solving instruction in this way has a number of benefits
+grounded in cognitive science research: first, it serves to activate
+prior knowledge\autocite{ambrose_how_2010}. With problem-solving
+strategies fresh in their memory, learners will more easily apply them
+when they begin programming. Second, it helps manage cognitive
+load. Where as in the previous example of language choice we reduced
+cognitive load by choosing a language with a simpler syntax, here we
+separate two tasks that we know individually demand a high level of
+cognitive load for novices: learning problem-solving strategies, and
+learning programming techniques.
+
+\section{Choice of problems to assign}
 As previously noted, OOP provides a set of tools that naturally lend
-themselves to solving certain types of problems, but not others. An
+themselves to solving certain types of problems but not others. An
 educator teaching OOP should carefully select problems and challenges
-that make sense to be solved with an OOP paradigm. Of course,
-provoiding illustrative examples of problems that are \emph{not} best
+that make sense to be solved with an OOP approach. Of course,
+providing illustrative examples of problems that are \emph{not} best
 solved with an OOP approach would be useful provided there is also
@@
 problems and brief mention of alternative approaches.
+
+\section{Assessment}
+Often times assessment becomes the weak point of a learning
+system. Designing good assessment is
+hard\autocite{lister_blooming_2000,traynor_synthesis_2005}, and often good assessment can
+require significant time on the part of the instructor. Because of
+this, assessment is often the weak point, lowering the effectiveness
+of the entire system.
+
+If our course objective is to teach learners to solve problems with
+programming then an ideal assessment would be to give students a
+problem to solve, ask them to write a program to solve it as well as
+reflect upon the patterns they used in their solution. Asking for
+reflection prompts students to engage in meta-cognition, becoming
+aware of their own thinking process, which helps them become more
+effective
+learners\autocite{ambrose_how_2010,chi_selfexplanations_1989}. Of
+course evaluating this type of assessment is much more time consuming
+than a multiple-choice exam that can be automatically scored, or even
+than evaluating that the finished program runs correctly, which can be
+done automatically as well\autocite{cheang_automated_2003}.
+
+\section{Considerations and Limitations}
+In the above hypothetical example we had the luxury of complete
+freedom in designing all aspects of an introductory programming
+curriculum. We recognize this is almost always never the case in
+practice. However, we argue that following a framework such as
+constructive alignment as close as possible will help. Thinking about
+the framework will also encourage critical engagement with the
+constraints imposed by the institution. For example, there has been
+much interest in assessment tools used for introductory programming
+courses\autocite{lister_blooming_2000,lister_objectives_2001,sheard_exploring_2011},
+and in particular automated
+assessment\autocite{cheang_automated_2003,traynor_synthesis_2005,ihantola_review_2010}. In
+many cases the push towards automated assessment techniques is a
+response to increasing class sizes that instructors typically have
+little control over. In the event that a large class size dictates
+that some form of automated assessment be used, following the
+constructive alignment framework would constrain the types of course
+objectives and instruction that were feasible. For instance, if the
+primary course objective was to teach conceptual understanding of
+computer science, then a well-designed multiple-choice assessment
+could be used
+effectively\autocite{lister_blooming_2000,lister_objectives_2001}.
+
+\section{Discussion and future work}
+
+This literature review was by no means comprehensive, but we hope that
+it touched upon some of the notable areas in which educators have been
+experiencing challenges and successes in teaching introductory
+programming. We have pointed out a number of possible connections
+between computer science and cognitive science that may be worth
+deeper investigation. While we have seen how cognitive load theory can
+help reason about course design, there are also potentially
+interesting connects between cognitive load theory and object oriented
+design: objects are designed to encapsulate complexity to better
+manage it. Additionally, further exploring concepts of knowledge
+organization and differences between novices and exports may be a
+fruitful area to investigate the use of teaching programming to aid in
+analogical reasoning and abstract
+thinking\autocite{bennedssen_abstraction_2008}.
+
+\section{Conclusion}
+
+Recognizing that introductory programming is a common component of
+engineering curriculum, and that there is a national push to introduce
+it to an even larger population, it is important to think carefully
+about how we design programming education to impart the cognitive
+skills that will generally benefit learners even after the specific
+technologies they learn are no longer relevant. We strongly advocate
+for the use a framework such as constructive alignment, described in
+this paper, to inform the choices for course objectives, instructional
+techniques, and assessments. Using such a framework well help identify
+components of a course that may be contributing unnecessarily to
+cognitive load, or reasons why student outcomes might not match
+expectations assumed from course objectives.
 
@@
 
-\section{Discussion}
+% \section{Future Work}
 
-\section{Future Work}
+% There continues to be a disconnect between learning sciences and how
+% programming paradigms and languages are selected for introductory
+% courses. While there is quite a bit of exploration around connections
+% between abstract thinking, threshold concepts, etc. and ``success'',
+% “success” is not consistently defined, sometimes meaning the ability
+% to write programs (e.g. with specifications given), sometimes focusing
+% more on the ability to solve problems.
 
-There continues to be a disconnect between learning sciences and how
-programming paradigms and languages are selected for introductory
-courses. While there is quite a bit of exploration around connections
-between abstract thinking, threshold concepts, etc. and ``success'',
-“success” is not consistently defined, sometimes meaning the ability
-to write programs (e.g. with specifications given), sometimes focusing
-more on the ability to solve problems.
+% \subsection{Continued Ties to Learning Sciences}
 
-\subsection{Continued Ties to Learning Sciences}
+% The thread of learning science influence split into two branches, one
+% further exploring the pattern-based approach and one exploring
+% threshold concepts. While developing problem-solving skills remains a
+% goal of CS1 only the pattern-based approach thread seems to focus on
+% this aspect while the researchers exploring threshold concepts are
+% more concerned with overall success in computer science without
+% specific mention of problem-solving skills.
 
-The thread of learning science influence split into two branches, one
-further exploring the pattern-based approach and one exploring
-threshold concepts. While developing problem-solving skills remains a
-goal of CS1 only the pattern-based approach thread seems to focus on
-this aspect while the researchers exploring threshold concepts are
-more concerned with overall success in computer science without
-specific mention of problem-solving skills.
+% \subsubsection{Problem Solving}
 
-\subsubsection{Problem Solving}
+% Several studies pick up the pattern-based focus of programming to
+% encourage problem-solving skill development from the
+% 1990s\autocite{muller_almost_2004,muller_pattern_2005,muller_patternoriented_2007,caspersen_instructional_2007-1}.
 
-Several studies pick up the pattern-based focus of programming to
-encourage problem-solving skill development from the
-1990s\autocite{muller_almost_2004,muller_pattern_2005,muller_patternoriented_2007,caspersen_instructional_2007-1}.
+% In one of these studies, the authors draw from three concepts of
+% cognitive science: cognitive load theory, cognitive apprenticeship,
+% and the theory of worked examples, to design an introductory
+% programming course\autocite{caspersen_instructional_2007-1}. It is
+% interesting to note that they select these theories based on a human
+% cognition model that resulted from computing: separate working and
+% long term memory where learning is the process for transforming
+% information from working to long-term memory. The course the authors
+% describe teachs OOP but they never explain why the chose OOP as the
+% paradigm, it is assumed that this is just a given feature of an
+% introductory programming course.
 
-In one of these studies, the authors draw from three concepts of
-cognitive science: cognitive load theory, cognitive apprenticeship,
-and the theory of worked examples, to design an introductory
-programming course\autocite{caspersen_instructional_2007-1}. It is
-interesting to note that they select these theories based on a human
-cognition model that resulted from computing: separate working and
-long term memory where learning is the process for transforming
-information from working to long-term memory. The course the authors
-describe teachs OOP but they never explain why the chose OOP as the
-paradigm, it is assumed that this is just a given feature of an
-introductory programming course.
+% %%TODO: consider focusing on this and dropping threshold concepts and others. Potentally bring in pattern-based pedagogy
 
-%%TODO: consider focusing on this and dropping threshold concepts and others. Potentally bring in pattern-based pedagogy
+% \subsubsection{Threshold Concepts}
 
-\subsubsection{Threshold Concepts}
+% Some studies explore potential threshold
+% topics\autocite{meyer_threshold_2005} for computer science and how
+% they might be useful for informing course
+% design\autocite{mccartney_liminal_2009,boustedt_threshold_2007}. \citeauthor{boustedt_threshold_2007}
+% identify object oriented programming and pointers as candidate
+% threshold topics based on interviews with students graduating with
+% computer science degrees. The two concepts and so these concepts are
+% in the context of computer science in general rather than introductory
+% programming, thus there is not a clear link between these concepts and
+% the goal of problem solving. In addition the authors note that OOP in
+% particular is quite broad and suggest further work to narrow the focus
+% of that threshold topic to one ore more specific aspects of OOP, such
+% as how objects communicate between one another.
 
-Some studies explore potential threshold
-topics\autocite{meyer_threshold_2005} for computer science and how
-they might be useful for informing course
-design\autocite{mccartney_liminal_2009,boustedt_threshold_2007}. \citeauthor{boustedt_threshold_2007}
-identify object oriented programming and pointers as candidate
-threshold topics based on interviews with students graduating with
-computer science degrees. The two concepts and so these concepts are
-in the context of computer science in general rather than introductory
-programming, thus there is not a clear link between these concepts and
-the goal of problem solving. In addition the authors note that OOP in
-particular is quite broad and suggest further work to narrow the focus
-of that threshold topic to one ore more specific aspects of OOP, such
-as how objects communicate between one another.
+% \subsubsection{Objects-first vs. X}
+% Much of the literature focuses on the debate between introducing
+% object oriented programming in CS1 or whether introductory courses
+% should teach a procedural style. Several researchers argue that an
+% object-first approach emphasices coding constructs over other goals
+% such as problem-solving and design
+% \autocite{moritz_objectsfirst_2005,muller_guidelines_2005,reges_back_2006}. \textcite{moritz_objectsfirst_2005}
+% propose a “design-first” approach that gets students to analyze
+% problems and design solutions before writing any computer code.
 
-\subsubsection{Objects-first vs. X}
-Much of the literature focuses on the debate between introducing
-object oriented programming in CS1 or whether introductory courses
-should teach a procedural style. Several researchers argue that an
-object-first approach emphasices coding constructs over other goals
-such as problem-solving and design
-\autocite{moritz_objectsfirst_2005,muller_guidelines_2005,reges_back_2006}. \textcite{moritz_objectsfirst_2005}
-propose a “design-first” approach that gets students to analyze
-problems and design solutions before writing any computer code.
+% \autocite{vilner_fundamental_2007} found no significant difference in
+% student performance on questions relating to recursion and efficiency
+% between groups that were given an object-first curriculum vs. groups
+% given a procedural-first curriculum. They did observe a significant
+% difference between the two groups when asked to design an abstract
+% data type: those who were in the objects-first group performed better
+% than those in the procedural-first group. However, it is notable that
+% the students were only tested on programming knowledge, not their
+% ability to analyze problems and design solutions.
 
-\autocite{vilner_fundamental_2007} found no significant difference in
-student performance on questions relating to recursion and efficiency
-between groups that were given an object-first curriculum vs. groups
-given a procedural-first curriculum. They did observe a significant
-difference between the two groups when asked to design an abstract
-data type: those who were in the objects-first group performed better
-than those in the procedural-first group. However, it is notable that
-the students were only tested on programming knowledge, not their
-ability to analyze problems and design solutions.
-
-\textcite{reges_back_2006} takes a ``back to basics'' approach eschewing
-the popular OOP and reverts his CS1 course back to a procedural style
-which allows him to put the emphasis back on problem solving.
+% \textcite{reges_back_2006} takes a ``back to basics'' approach eschewing
+% the popular OOP and reverts his CS1 course back to a procedural style
+% which allows him to put the emphasis back on problem solving.
 

### 2016-02-03 paper.tex
Updates from ShareLaTeX

--- 2016-02-02 paper.tex
@@
 \documentclass[12pt]{article}
+
+\usepackage[english]{babel}
+\usepackage[utf8]{inputenc}
+\usepackage[babel]{csquotes}
+\usepackage{graphicx}
+\usepackage{etoolbox}
+\usepackage{changepage}
+\usepackage{titlesec}
+\usepackage{hyperref}
+\usepackage[parfill]{parskip}
+\usepackage[margin=1in]{geometry}
+
+%\usepackage{xunicode}
+\usepackage[dvipsnames,table]{xcolor}
+\usepackage{tikz}
+\usetikzlibrary{positioning,arrows,shapes,fit}
+\usepackage{standalone}
 
@@
 \usepackage[backend=biber,autolang=hyphen, bibencoding=inputenc, isbn=false, uniquename=false, sorting=none, sortcites=true, style=numeric, autocite=superscript, url=false]{biblatex} % biblatex setup
-\DeclareLanguageMapping{american}{american-apa}
+\addbibresource{./computing.bib}
 \appto{\bibfont}{\footnotesize} % for 10pt bibliography
-%\usepackage{xunicode}
-\usepackage[american]{babel}
-\usepackage[babel]{csquotes}
-\usepackage[utf8]{inputenc}
-\usepackage{float}
-\usepackage{graphicx}
-\usepackage[margin=1in]{geometry}
-\usepackage[dvipsnames,table]{xcolor}
-\usepackage{tikz}
-\usetikzlibrary{positioning,arrows,shapes,fit}
-\usepackage{standalone}
-%\usepackage{mdframed}
-\usepackage{changepage}
-\usepackage{titlesec}
-\usepackage[parfill]{parskip}
 
@@
 \let\textcite=\autocite
-\addbibresource{computing.bib}
 
@@
 % conflict between teaching programming syntax, teaching ``how to program'', teaching high-level concepts like abstractions and how to ``do'' problem solving.
-\section{Coding for Everyone}
+\section*{Coding for Everyone}
 In the past few years, there has been increasing interest to get more

### 2016-02-03 paper.tex

--- 2016-02-03 paper.tex
@@
 \usepackage[backend=biber,autolang=hyphen, bibencoding=inputenc, isbn=false, uniquename=false, sorting=none, sortcites=true, style=numeric, autocite=superscript, url=false]{biblatex} % biblatex setup
-\DeclareLanguageMapping{american}{american-apa}
+\addbibresource{./computing.bib}
 \appto{\bibfont}{\footnotesize} % for 10pt bibliography
-%\usepackage{xunicode}
-\usepackage[american]{babel}
-\usepackage[babel]{csquotes}
-\usepackage[utf8]{inputenc}
-\usepackage{float}
-\usepackage{graphicx}
-\usepackage[margin=1in]{geometry}
-\usepackage[dvipsnames,table]{xcolor}
-\usepackage{tikz}
-\usetikzlibrary{positioning,arrows,shapes,fit}
-\usepackage{standalone}
-%\usepackage{mdframed}
-\usepackage{changepage}
-\usepackage{titlesec}
-\usepackage[parfill]{parskip}
-\usepackage{mdframed}
 
@@
 
-%%\let\textcite=\autocite
-\addbibresource{computing.bib}
+\let\textcite=\autocite
 
@@
 
-\subsection{Problem Solving}
+\subsection{Problem solving}
 Developing problem solving was widely regarded as a common objective
@@
 
-\section{Considerations and Limitations}
+\section{Considerations and limitations}
 In the above hypothetical example we had the luxury of complete

### 2016-02-04 paper.tex
added example patterns for procedural and oop

@@
 programming\autocite{robins_learning_2003}. They found that the
-biggest challenge for novice programmers was not understanding the
+biggest challenge for novice programmers was not in understanding the
 basic concepts, but knowing how to apply them. This finding is
-consistent understanding of how we
+consistent with current understanding of how we
 learn\autocite{bransford_how_2000,ambrose_chapter_2010}, and with
@@
 suggestion is that helping students learn strategies that work is
-crucial to helping them become effective novices, i.e. devote
+crucial to helping them become \emph{effective} novices, i.e. devote
 instructional time to problem-solving strategies and how to apply
@@
 when to apply them to particular
-problems\autocite{east_patternbased_1996,muller_pattern_2005,wallingford_first_1996}.
+problems\autocite{east_patternbased_1996,muller_pattern_2005,wallingford_first_1996,reed_incorporating_1998}.
 
@@
 systems\autocite{kay_history_1996}. The underlying metaphore of Kay's
-conception of OOP was that of the biological cell: each object is
+conception for OOP was the biological cell: each object is
 designed to accomplish a specific set of tasks where the complexity of
@@
 set of tools for problem solving, and it is these tools and concepts
-that become the
-``basics''\autocite{kolling_problem_1999,kay_history_1996}.
+that become the ``basics'': objects, messages, inheritance, and
+polymorphism\autocite{kolling_problem_1999,kay_history_1996}.
 
 For instance, if taking a patterns-based approach to teaching
-programming one could use either a procedural paradigm or an OOP
+programming, one could use either a procedural paradigm or an OOP
 paradigm, but the fundamental patterns would be quite different
-depending on the paradigm\autocite{wallingford_first_1996}.
+depending on the paradigm\autocite{wallingford_first_1996}. Useful
+procedural patterns might be:
+
+\begin{description}
+\item[read-evaluate-print] reading a piece of data, evaluating it, and output the result
+\item[guarded-action] if a guard-condition is satisfied, take action.
+\end{description}
+
+While useful OOP patterns might be:
+
+\begin{description}
+\item[state]  provide controlled access to a body of data
+\item[view] decouple an object's state from its representation
+\item[decorator] add functionally to an object without modifying its internal structure.
+\end{description}
 

### 2016-02-04 paper.tex
Updates from ShareLaTeX

--- 2016-02-03 paper.tex
@@
 \documentclass[12pt]{article}
-
 \usepackage[english]{babel}
@@
 \usepackage[margin=1in]{geometry}
-
+\usepackage{times}
 %\usepackage{xunicode}
@@
 \usepackage{standalone}
+\usepackage{textcomp}
 
@@
 % conflict between teaching programming syntax, teaching ``how to program'', teaching high-level concepts like abstractions and how to ``do'' problem solving.
-\section{Coding for everyone}
+\section*{Coding for everyone}
 In the past few years, there has been increasing interest to get more
@@
 
-\section{Methods}
+\section*{Methods}
 Assuming that there would be more comprehensive research of

### 2016-03-21 paper.tex
grammatical fixes

@@
 introductory programming in the computer science community we began
-with targeted searchs of the ACM digital library for `introductory
+with targeted searches of the ACM digital library for `introductory
 programming'. Once we found key papers that seemed to indicate a call
@@
 making a popular practice part of a learning environment without
-critically analyzing whether or not is a good fit is by no means
+critically analyzing whether or not is a good fit, is by no means
 exclusive to programming education. For example, flipped
@@
 practitioners flipping their classroom without adjusting course
-objectives, content and activities to work under the flipped-classroom
+objectives, content, and activities to work under the flipped-classroom
 model\autocite{desantis_students_2015}. Likewise, anyone can teach
@@
 object oriented programming should be placed within the curriculum
-with debates between objects-first vs procedural-fist but little
+with debates between objects-first vs procedural-fist, but little
 question of why object oriented design should be included at all other
-than it continues to be a popular paradigm in industry.
+than that it continues to be a popular paradigm in industry.
 
@@
 \item Even if goals were clearly stated it is likely that many
-  different approaches could be used successful if chosen
-  intentionally to align with the goal and measurements of success
-  (assessment). In many of the studies it is not immediately clear if
+  different approaches could be used successfully if chosen
+  intentionally to align with the goals and measurements of success
+  (assessment). In many of the studies, it is not immediately clear if
   the measures of success used are appropriate for the stated
@@
 system}\label{sec:components-of-a-learning-system}
-
+%% Reviewer: short
 A popular way to visualize an effective learning system is as a triad
@@
 \subsection{Object oriented programming}
-
+%% Reviewer: more explicit connections to cogsci
 Alan Kay originally conceived of object oriented programming as
@@
 paradigm. Unsurprisingly, there is no paradigm that is the ``best''
-fit for every problem. Leaving students with the believe that only one
+fit for every problem. Leaving students with the belief that only one
 paradigm exists makes it very difficult for them to learn how to
@@
 taught \emph{first} since the reverse shift, from OOP to procedural,
-is conceptually easier.
+is conceptually easier. What this
 
@@
 \section{Choice of problems to assign}
+%% Reviewer: short
 As previously noted, OOP provides a set of tools that naturally lend
@@
 \section{Assessment}
+%% Reviewer: short
 Often times assessment becomes the weak point of a learning
@@
 curriculum. We recognize this is almost always never the case in
-practice. However, we argue that following a framework such as
-constructive alignment as close as possible will help. Thinking about
+practice; however, we argue that following a framework, such as
+constructive alignment, as close as possible will help. Thinking about
 the framework will also encourage critical engagement with the
@@
 technologies they learn are no longer relevant. We strongly advocate
-for the use a framework such as constructive alignment, described in
+for the use of a framework such as constructive alignment, described in
 this paper, to inform the choices for course objectives, instructional
@@
 
+\subsection{Cognitive Load Theory}
+
+\section{Notes for Revisions}
+
+If teaching OOP concepts is a goal of the course one must ask if
+/implementation/ of OOP programs is also a goal.  If writing OOP code
+is a goal and it is unreasonable to assume that students will have
+already mastered syntax and enough understanding of the development
+environment (interpreting compiler messages, debugging, etc) then CLT
+would suggest that extensive /scaffolding/ be employed to reduce
+extraneous cognitive load. Being careful to sequence simple-to-complex
+concepts will help reduce /intrinsic/ cognitive load: i.e. teach bare
+objects and make sure students have began to automate tasks associated
+with implementing bare objects before introducing inheritance.
+
+Keep in mind that providing scaffolding for syntax and grammar,
+e.g. in the form of worked examples, or implementation in which
+students need to copy/paste implementation code into object template
+code may help with teaching OOP concepts, but would not be aiding
+learning of actually implementing OOP code from scratch. Simply seeing
+the implemented code is likely not enough to aid most students in
+learning this skill themselves and if that is a goal of the course, or
+a subsequent course it is important to focus on that while scaffolding
+other aspects (e.g. providing an object interface and description of
+how the object is intended to function, ask students to write
+implementation). The main challenge with implementing such an
+instructional strategy is that it will likely require coordination
+across courses and instructors as teaching both programming, OOP,
+debugging strategies all at once will likely not be met with much
+success. A possible exception might be if one considers a spiral
+curriculum, but the success of this still depends on close
+coordination across courses and instructors.
+
+The authors of the Lohani et al report describe efforts to use the
+Alice programing environment to teach introductory programming
+concepts\autocite{lohani_reformulating_2011}. In theory a graphical
+programming environment should aid in the , learning of higher level
+concepts since it eliminates the extraneous cognitive load of worrying
+about syntax and grammar. In fact, there has been success in using
+iconic languages to teach first year programming
+\autocite{calloni_iconic_1997}, and research indicates that teaching
+graphical programming first does not inhibit a student from later
+learning textual programming languages\autocite{hundhausen_can_2009}.
+
+Eric Raymond describes three different kinds of complexity in software
+design: essential complexity, optional complexity, and accidental
+complexity
+
 \vspace{4\baselineskip}\vspace{-\parskip} % Creaters proper 4 blank line spacing.

### 2016-03-21 paper.tex
expand CLT section, connect to OOP

@@
 \titleformat{\section}{\normalsize\bfseries}{}{}{}[]
-\titleformat{\subsection}{\bfseries}{}{}{}[]
-\titleformat{\subsubsection}{\bfseries}{}{}{}[]
+\titleformat{\subsection}[runin]{\bfseries}{}{}{}[.]
+\titleformat{\subsubsection}[runin]{\itshape}{}{}{}[.]
 
@@
 taught \emph{first} since the reverse shift, from OOP to procedural,
-is conceptually easier. What this
+is conceptually easier. What this recommendation neglects to account
+for however is that both paradigms depend upon authoring programs,
+whether they be textual or iconic, and this authoring process takes a
+certain amount of cognitive capacity.
+
+\subsection{Cognitive load theory}
+Cognitive load theory (CLT) is based on the premise that the capacity
+of Working Memory, where all conscious cognitive processing takes
+place, has a finite and relatively small capacity: we are able to
+manage 2-3 interacting elements at a time in our working
+memory. Examples of ``interacting elements'' in the current context
+would be something like recalling and using the syntax and grammar of
+a particular programming language, reasoning about the behavior of a
+single object in an OOP setting, or reasoning about a solution to a
+particular high-level problem. CLT posits that learners have a finite
+capacity for their short-term memory, and for learning to occur, all
+aspects of the task must fit into this finite space. While some
+contributors to cognitive load are necessary for the task at hand,
+others are not\autocite{paas_cognitive_2003}.
+
+In a superficial way, CLT theory would suggest that OOP is a helpful
+paradigm for managing complexity: objects encapsulate complexity by
+hiding the details of implementation behind a simple interface. For
+example, and object that represents a list of items may have
+operations to add and remove items from the list. Conceptually we can
+use this object with our common understanding of what it means to
+``add'' and ``item'' to a ``list''. Critically, we do \emph{not} need
+to be aware of how the adding and removal of the item is actually
+implemented in code, or even whether how the storage is structured in
+memory (e.g. an array or a linked-list). We refer to this use of CLT
+to justify the reduction of complexity of OOP as superficial because
+it focus on the conceptual paradigm itself without regard to
+implementing it in some programming environment. In practice, we
+usually ask our students to implement OOP concepts in a program. That
+task requires not only understanding OOP as a paradigm, but also
+working with the programming environment, authoring a program, and
+most likely debugging it as well.
 
@@
 complex syntax features\autocite{koulouri_teaching_2014}. The
-explanation for this can be found in cognitive load theory (CLT),
-which posits that learners have a finite capacity for their short-term
-memory, and for learning to occur, all aspects of the task must fit
-into this finite space. While some contributors to cognitive load are
-necessary for the task at hand, others are
-not\autocite{paas_cognitive_2003}. If, for example, the goal is to
+explanation for this can be found in CLT. If, for example, the goal is to
 teach problem-solving techniques, there will be cognitive load
@@
 the language\autocite{goldwasser_teaching_2008}.
+
+\subsubsection{Iconic versus textual languages}
+
+In theory a iconic programming environment such as
+Scratch\autocite{resnick_scratch_2009} or
+Alice\autocite{cooper_using_2003} should aid in the learning of higher
+level concepts since it eliminates the extraneous cognitive load of
+dealing with syntax and grammar of a language. In fact, there has been
+success in using iconic languages to teach first year programming
+\autocite{calloni_iconic_1997}, and research indicates that teaching
+graphical programming first does not inhibit a student from later
+learning textual programming languages\autocite{hundhausen_can_2009}.
+
+\citeauthor{snook_incorporation_2005} describe efforts to use the Alice
+programing environment to teach introductory programming
+concepts\autocite{snook_incorporation_2005}. Interestingly a later
+report assessing the efficacy of the curriculum which included this
+use of Alice indicated that while pre-post testing indicated learning
+gains with Alice, focus group data indicated a dissatisfaction with
+Alice as an introductory language which was later replaced with
+LabVIEW\autocite{lohani_reformulating_2011}.
 
@@
 
-\subsection{Cognitive Load Theory}
-
 \section{Notes for Revisions}
@@
 If teaching OOP concepts is a goal of the course one must ask if
-/implementation/ of OOP programs is also a goal.  If writing OOP code
-is a goal and it is unreasonable to assume that students will have
-already mastered syntax and enough understanding of the development
-environment (interpreting compiler messages, debugging, etc) then CLT
-would suggest that extensive /scaffolding/ be employed to reduce
-extraneous cognitive load. Being careful to sequence simple-to-complex
-concepts will help reduce /intrinsic/ cognitive load: i.e. teach bare
-objects and make sure students have began to automate tasks associated
-with implementing bare objects before introducing inheritance.
+\emph{implementation} of OOP programs is also a goal.  If writing OOP
+code is a goal and it is unreasonable to assume that students will
+have already mastered syntax and enough understanding of the
+development environment (interpreting compiler messages, debugging,
+etc) then CLT would suggest that extensive \emph{scaffolding} be
+employed to reduce extraneous cognitive load. Being careful to
+sequence simple-to-complex concepts will help reduce /intrinsic/
+cognitive load: i.e. teach bare objects and make sure students have
+began to automate tasks associated with implementing bare objects
+before introducing inheritance.
 
@@
 
-The authors of the Lohani et al report describe efforts to use the
-Alice programing environment to teach introductory programming
-concepts\autocite{lohani_reformulating_2011}. In theory a graphical
-programming environment should aid in the , learning of higher level
-concepts since it eliminates the extraneous cognitive load of worrying
-about syntax and grammar. In fact, there has been success in using
-iconic languages to teach first year programming
-\autocite{calloni_iconic_1997}, and research indicates that teaching
-graphical programming first does not inhibit a student from later
-learning textual programming languages\autocite{hundhausen_can_2009}.
-
 Eric Raymond describes three different kinds of complexity in software
 design: essential complexity, optional complexity, and accidental
-complexity
+complexity. This has interesting parallels to the different types of
+cognitive load in CLT. In particular the notion of \emph{essential
+  complexity}, like \emph{intrinsic} cognitive load is associated with
+the the solution itself: a solution to a complex problem will be
+essentially complex due to the complexity associated with the problem
+itself.
 

### 2016-03-22 paper.tex
final edits

--- 2016-03-21 paper.tex
@@
 \section*{Coding for everyone}
-In the past few years, there has been increasing interest to get more
-people, regardless of background, to learn to program. In fact,
-President Obama recently made a statement calling on all children to
-have the opportunity to learn about computer
+In the past few years, there has been increasing interest in
+encouraging more people, regardless of background, to learn to
+program. In fact, President Obama recently made a statement calling on
+all children to have the opportunity to learn about computer
 science\autocite{whitehouse_computer_2016}. Sites such as code.org
 promote CS education opportunities for all, citing statistics about
-STEM jobs and arguing that in the 21st century knowledge about
+STEM jobs and arguing that in the 21st century, knowledge about
 computer science is foundational\autocite{code.org_every_2016}. Many
@@
 in general ``basic programming skills'' is a common desired outcome
-across engineering
-curricula\autocite{reid_classification_2014}.
+across engineering curricula\autocite{reid_classification_2014}.
 
@@
 technology is changing rapidly, and the half-life of technical skills
-is decreasing\autocite{nae_educating_2004}, it would seem that
+is decreasing\autocite{nae_educating_2004}, might seem to indicate that
 teaching programming is a questionable use of class time: the language
@@
 acknowledge that learning to program is more than simply learning to
-write code, and in particular if we draw on work from the early era of
+write code, and in particular, if we draw on work from the early era of
 computing which closely linked programming to problem solving and
@@
 by this was that computing allows for new ways of dealing with
-complexity, provides new metaphors for thinking about
+complexity, and provides new metaphors for thinking about
 systems\autocite{kay_history_1996}. Early work demonstrated that with
@@
 graduate may be beside the point, there are lasting cognitive benefits
-learning to interact with a computer.
+learning to interact with a computer can provide.
 
@@
 In this paper we will present a brief summary of current published
-work on introductory programming education. In particular we will
+work on introductory programming education. In particular, we will
 explore the alignment of course objectives, pedagogical strategies,
@@
 about both the design and evaluation of introductory programming
-courses.
+courses. We will then provide a hypothetical case study of a
+introductory programming course organized around this framework and
+discuss implications of different instructional strategies through the
+lens of a theory of cognitive processes called cognitive load theory.
 
@@
 Assuming that there would be more comprehensive research of
-introductory programming in the computer science community we began
-with targeted searches of the ACM digital library for `introductory
-programming'. Once we found key papers that seemed to indicate a call
-to action
+introductory programming in the computer science community, we began
+with targeted searches of the ACM digital library for the key phrase
+`introductory programming'. Once we found key papers that seemed to
+indicate a call to action
 \autocite[e.g.][]{mccracken_multinational_2001,kolling_problem_1999},
@@
 
-Because the literature search of introductory programming seemed to be
-interest in why and how programming was taught, we needed a documented
-framework to consistently evaluate and starting from
+Because the literature search of introductory programming resulted in
+variety of examples of how and why programming was taught, we needed a
+documented framework to consistently evaluate each one and guide our
+hypothetical course design. Starting from
 \citeauthor{jamieson_creating_2009}'s
-(\citeyear{jamieson_creating_2009}) to ground educational research
-activities in how people learn we identified the concept of
-\emph{constructive alignment} as a promising one. Following the
-references in \citeauthor{jamieson_creating_2009}'s report as well as
-conducting targeted literature searches for ``constructive alignment''
-we were able to find descriptions of this
+(\citeyear{jamieson_creating_2009}) call to ground educational
+research activities in how people learn we identified the concept of
+\emph{constructive alignment} as a promising framework to evaluate and
+design learning environments. Following the references in
+\citeauthor{jamieson_creating_2009}'s report as well as conducting
+targeted literature searches for ``constructive alignment'' we were
+able to find descriptions of this
 framework\autocite{jamieson_creating_2009,pellegrino_rethinking_2006,felder_designing_2003,biggs_enhancing_1996}.
@@
 vs. Java\autocite{goldwasser_teaching_2008,koulouri_teaching_2014}),
-and best teaching practices (e.g. use of “pair programming”
+and best teaching practices (e.g. use of ``pair programming''
 environments\autocite{williams_support_2001,mcdowell_effects_2002,thomas_code_2003,chong_social_2007,salleh_empirical_2011}). However,
@@
 practitioners flipping their classroom without adjusting course
-objectives, content, and activities to work under the flipped-classroom
-model\autocite{desantis_students_2015}. Likewise, anyone can teach
-elements of object oriented programming in their introductory
-programming courses, but unless critical aspects of the course are
-adjusted: background reading, rethinking what the ``fundamental
-concepts'' are, the types of problems assigned, and methods of
-assessment, the incorporation of OOP will likely create confusion for
-students and frustration for instructor\autocite{kolling_problem_1999}.
+objectives, content, and activities to work under the
+flipped-classroom model\autocite{desantis_students_2015}. Likewise,
+anyone can teach elements of object oriented programming in their
+introductory programming courses, but unless critical aspects of the
+course are adjusted: background reading, rethinking what the
+``fundamental concepts'' are, the types of problems assigned, and
+methods of assessment, the incorporation of OOP will likely create
+confusion for students and frustration for
+instructor\autocite{kolling_problem_1999}.
 
@@
 While the \citeauthor{mccracken_multinational_2001} study validated the
-experiences of many introductory programming instructors it did not
+experiences of many introductory programming instructors, it did not
 result in critical change to address the problems. Rather, what we
-have seen since the 1990s is been an increasing interest in where
+have seen since the 1990s has been an increasing interest in where
 object oriented programming should be placed within the curriculum
@@
 There remains a desire that introductory programming courses should
-teach problem-solving skills, though only a few papers present
+teach problem-solving skills, though only a few authors present
 evidence that the implementation of this goal is informed by cognitive
@@
 that despite the volume of work exploring the implementation of
-introductory programming courses “there is little systematic evidence
-to support any particular approach.”\autocite{pears_survey_2007-1}
+introductory programming courses ``there is little systematic evidence
+to support any particular approach.''\autocite{pears_survey_2007-1}
 We think this could be due to two reasons:
@@
   computing and computer science, or the ability to analyze problems
-  and design solutions (problem-solving). The “appropriate” approach
+  and design solutions (problem-solving). The ``appropriate'' approach
   to teaching would depend on the goal.
@@
 language with a simpler syntax (Python instead of Java), and
-introducing explicit problem-solving training improved student
-learning, while providing more formative assessment did not seem to
-help. They speculate that the reason for the latter finding could be
-that students do not spend time looking at formative feedback, or do
-not understand what is given.
+introducing explicit problem-solving training, improved student
+learning. They also found, however, that providing more formative
+assessment did not seem to help. They speculate that the reason for
+the latter finding could be that students do not spend time looking at
+formative feedback, or do not understand the feedback that is given.
 
@@
 example, if an objective of a programming course is that learners will
-develop problem solving skills, but assessment only consist of test of
-programming concepts and syntax, teaching/learning activities would
+develop problem solving skills, but assessment only consists of tests
+of programming concepts and syntax, teaching/learning activities would
 tend to center around those that promote understanding of these
@@
 ``best'' teaching paradigm. Rather, we want to point out that from
-what has been reported teaching OOP effectively can be
+what has been reported, teaching OOP effectively can be
 challenging. What many implementations seem to miss is that OOP
@@
 objects\autocite{raymond_art_2003}. Different problems may be better
-solved using a procedural, or functional
+solved using a procedural, functional, or data-flow
 paradigm\autocite{kay_history_1996,felleisen_functional_2009,crestani_experience_2010}. Ultimately,
@@
 sets\autocite{kolling_problem_1999,vilner_fundamental_2007}.
+
+\subsection{Cognitive load theory}
+Cognitive load theory (CLT) is based on the premise that the capacity
+of our working memory, where all conscious cognitive processing takes
+place, is finite and relatively small: we are able to manage 2-3
+interacting elements at a time. Examples of ``interacting elements''
+in the current context would be recalling and using the syntax and
+grammar of a particular programming language, reasoning about the
+behavior of a single object in an OOP environment, or reasoning about
+a solution to a particular high-level problem. CLT posits that in
+light of this finite capacity of working memory, for learning to
+occur, all aspects of a task must fit into this finite space. While
+some contributors to cognitive load are necessary for the task at
+hand, others are not\autocite{paas_cognitive_2003}. Cognitive load can
+be divided into three different categories: \emph{intrinsic} cognitive
+load is due to the complexity of the task itself and generally can not
+be reduced without simplifying the task. \emph{Extraneous} cognitive
+load is associated with work that is due to the instructional method,
+but not part of the task at hand. For instance, if the educational
+goal is to complete a programming task, but learners must search and
+interpret reference material in order to do so, the search and
+interpretation of this material would be considered extraneous
+cognitive load. Finally, \emph{germane} cognitive load is also a
+result of the instructional design, but unlike extraneous load,
+germane load enhances the learning process. For instance, if a
+learning goal is that students are able to navigate and use reference
+material, than the previous example would be considered germane
+cognitive load. What this example highlights is that whether
+complexity due to the instruction design is extraneous or germane
+depends on the specific learning goals. The role of the instructional
+designer then is to reduce extraneous load and intentionally managing
+germane load.
 
@@
 programming'', usually defined as conditional statements, loops, and
-in some cases pointers. This concern is misplaced, under an OOP
+in some cases pointers. This concern is misplaced; under an OOP
 paradigm the ``basics of programming'' are not the same as when
@@
 Ultimately, if our goal is to teach problem solving in general, we
-would be doing a grave disservice to our students to leave with the
-impression that there is only one paradigm, whether it be OOP,
+would be doing a grave disservice to our students to leave them with
+the impression that there is only one paradigm, whether it be OOP,
 procedural, functional, or something not yet invented. As previously
@@
 taught \emph{first} since the reverse shift, from OOP to procedural,
-is conceptually easier. What this
+is conceptually easier. What this recommendation neglects to account
+for however is that both paradigms depend upon authoring programs in
+some language, and this authoring process takes a certain amount of
+cognitive capacity. To justify an objects-first approach we would have
+to verify that the cognitive load associated with OOP concepts
+themselves combined with the cognitive load of learning to author a
+program in a particular environment, do not exceed the learner's
+working memory capacity.
+
+In a superficial way, CLT theory would suggest that OOP is a helpful
+paradigm for managing cognitive omplexity: objects encapsulate
+complexity by hiding the details of implementation behind a simple
+interface. For example, an object that represents a list of items may
+have operations to add and remove items from the list. Conceptually we
+can use this object with our common understanding of what it means to
+``add'' an ``item'' to a ``list''. Critically, we do \emph{not} need
+to be aware of how the adding and removal of the item is actually
+implemented in code, or even how the storage is structured in memory
+(e.g. an array or a linked-list). We refer to this use of CLT to
+justify the reduction of complexity of OOP as superficial because it
+focuses on only the conceptual paradigm itself without regard to
+implementing it in some programming environment. In practice, we
+usually ask our students to implement OOP concepts in a particular
+programming language. That task requires not only understanding OOP as
+a paradigm, but also working with the programming environment,
+authoring a program, understanding the syntax and grammar of the
+language, and most likely debugging it as well.
+
+\subsection{Cognitive challenges teaching OOP}
+A behavior that characterizes expert programmers is the ability to
+easily shift between different abstraction layers of a problem
+space\autocite{armoni_teaching_2013}. This ability may be especially
+crucial to becoming fluent in OOP as the concepts are somewhat further
+removed from the actual syntax than in other paradigms. For example,
+in the procedural paradigm the central concept is that of the
+procedure, or subroutine. Subroutines are simply collections of
+imperative commands grouped together under a label. Thus, if someone
+can write and understand imperative code, the conceptual leap to
+understanding the procedural paradigm is relatively small. In
+contrast, to understand objects in OOP one must understand the concept
+of state and methods operating on the state. Both state and method are
+implemented with code (generally imperative), but to understand how an
+object functions one has to imagine its initial state, and then how
+each method call would change this state. Cognitive load theory would
+suggest that understanding the dynamic interaction of state and method
+calls is more challenging than grouping lines of imperative codes into
+re-usable procedures. Thus to be fluent in OOP design a programmer
+\emph{must} be able to switch between thinking at the object layer and
+the program layer. Teaching abstraction is best done by doing it by
+example, while being explicit about the abstractions
+used\autocite{armoni_teaching_2013}. However, we must keep in mind the
+limited capacity of working memory: if the cognitive load associated
+with writing code and thinking about state and methods fills up a
+learners working memory they may have difficult learning concepts of
+abstraction, even when the instructor explicitly points them out. To
+reduce the cognitive load associated with writing code and working
+with object state and methods, these tasks must be practiced until
+somewhat automated: they can be done without conscious
+thought\autocite{ambrose_chapter_2010}.
 
@@
 complex syntax features\autocite{koulouri_teaching_2014}. The
-explanation for this can be found in cognitive load theory (CLT),
-which posits that learners have a finite capacity for their short-term
-memory, and for learning to occur, all aspects of the task must fit
-into this finite space. While some contributors to cognitive load are
-necessary for the task at hand, others are
-not\autocite{paas_cognitive_2003}. If, for example, the goal is to
-teach problem-solving techniques, there will be cognitive load
+explanation for this can be found in CLT. If, for example, the goal is
+to teach problem-solving techniques, there will be cognitive load
 associated with thinking about different strategies one could
@@
 cognitive load, but it does not aid in the process of learning
-problem-solving techniques. Thus, it is in the educator's best
-interest to minimize cognitive load associated with features that do
-not directly help with the learning objectives of the course and
-select a teaching language with easy-to-learn syntax.
+problem-solving techniques. In other words, learning the syntax of a
+programming language is an extraneous cognitive load in this
+context. Thus, it is in the educator's best interest to minimize
+cognitive load associated with features that do not directly help with
+the learning objectives of the course and select a teaching language
+with easy-to-learn syntax.
 
 In an attempt to address the concern that the two primary languages,
-C++ and Java, used to teach OOP in introductory courses, may be too
+C++ and Java, used to teach OOP in introductory courses may be too
 complex for novice students, \citeauthor{goldwasser_teaching_2008}
@@
 the language\autocite{goldwasser_teaching_2008}.
+
+\subsubsection{Iconic versus textual languages}
+
+In theory an iconic programming environment such as
+Scratch\autocite{resnick_scratch_2009} or
+Alice\autocite{cooper_using_2003} should aid in the learning of higher
+level concepts, such as those associated with OOP, since graphical
+languages eliminates the extraneous cognitive load of dealing with
+syntax and grammar of a language. In fact, there has been success in
+using iconic languages to teach first year programming
+\autocite{calloni_iconic_1997}, and research indicates that teaching
+graphical programming first does not inhibit a student from later
+transferring learned high level concepts to textual programming
+languages\autocite{hundhausen_can_2009}.
+
+\citeauthor{snook_incorporation_2005} describe efforts to use the Alice
+programming environment to teach introductory programming
+concepts\autocite{snook_incorporation_2005}. Interestingly a later
+report assessing the efficacy of the curriculum using Alice indicated
+that while pre-post testing indicated learning gains with the
+environment, focus group data indicated a dissatisfaction with Alice
+as an introductory language resulting in a switch to
+LabVIEW\autocite{lohani_reformulating_2011}.
 
@@
 
-\section{Choice of problems to assign}
-%% Reviewer: short
-As previously noted, OOP provides a set of tools that naturally lend
-themselves to solving certain types of problems but not others. An
-educator teaching OOP should carefully select problems and challenges
-that make sense to be solved with an OOP approach. Of course,
-providing illustrative examples of problems that are \emph{not} best
-solved with an OOP approach would be useful provided there is also
-time for discussion of \emph{why} OOP might not be suited for certain
-problems and brief mention of alternative approaches.
+% \section{Choice of problems to assign}
+% %% Reviewer: short
+% As previously noted, OOP provides a set of tools that naturally lend
+% themselves to solving certain types of problems but not others. An
+% educator teaching OOP should carefully select problems and challenges
+% that make sense to be solved with an OOP approach. Of course,
+% providing illustrative examples of problems that are \emph{not} best
+% solved with an OOP approach would be useful provided there is also
+% time for discussion of \emph{why} OOP might not be suited for certain
+% problems and brief mention of alternative approaches.
 
@@
 
+An area that has not been addressed in this paper but has strong
+connections is that of attention in
+cognition\autocite{simon_role_1986}. There are strong parallels
+between our limitations to focus our conscious attention at one or two
+tasks at a time, and the serial processing inherent in most computing
+and language design. Of particular interest is the move in hardware to
+increasingly parallel designs to keep pace with Moore's law. Our
+cognitive constraints of attention may suggest an inherent difficulty
+in thinking and reasoning about parallel processing that must be
+addressed if we want to prepare students for an environment that
+demands parallel-processing solutions to programming problems. These
+demands may shift focus to languages such as Erlang that are designed
+for high levels of concurrency as well as demand we adopt different
+metaphors for computing\autocite{stein_challenging_1999}.
+
 \section{Conclusion}
@@
 expectations assumed from course objectives.
+
+Analyzing a proposed curriculum through the lens of cognitive load
+theory can aid in instruction decisions that will increase the chances
+achieving learning goals. Educators must critically review the goals
+of their introductory programming courses: if goals are to understand
+high level concepts such as OOP, but not necessarily implement these
+concepts in textual languages, it may be prudent to consider iconic
+languages for these courses. If however, goals include learning OOP
+concepts \emph{and} implementing these concepts in textual language,
+educators must be aware of the increased cognitive load associated
+with each of these tasks and provide scaffolding to help students
+automate aspects of each task individually before expecting them to
+combine them both in successful programs.
+
+Taken together, using an established framework, such as constructive
+alignment, to design an analyze curriculum, and utilizing cognitive
+load theory to help reduce extraneous cognitive load while being
+intentional about inclusion of germane cognitive load should help
+instructional designers create learning environments in which students
+are able to achieve learning goals. In some cases, intended goals may
+have to be adjusted to account for the finite limitations of working
+memory capacity: it is likely not feasible to expect students with no
+programming experience to learn the syntax and grammar of a new
+language \emph{and} higher level concepts associated with a paradigm
+like OOP, \emph{and} be able to switch between abstraction layers in
+the fluid manner necessary to solving complex problems with
+programs. What this suggests is that closer coordination is needed
+across courses in a curriculum so that goals of one become background
+knowledge of subsequent courses, resulting in successful achievement
+of program learning goals by the time of graduation.
 
@@
 
-\subsection{Cognitive Load Theory}
+% \section{Notes for Revisions}
 
-\section{Notes for Revisions}
+% If teaching OOP concepts is a goal of the course one must ask if
+% \emph{implementation} of OOP programs is also a goal.  If writing OOP
+% code is a goal and it is unreasonable to assume that students will
+% have already mastered syntax and enough understanding of the
+% development environment (interpreting compiler messages, debugging,
+% etc) then CLT would suggest that extensive \emph{scaffolding} be
+% employed to reduce extraneous cognitive load. Being careful to
+% sequence simple-to-complex concepts will help reduce \emph{intrinsic}
+% cognitive load: i.e. teach bare objects and make sure students have
+% began to automate tasks associated with implementing bare objects
+% before introducing inheritance.
 
-If teaching OOP concepts is a goal of the course one must ask if
-/implementation/ of OOP programs is also a goal.  If writing OOP code
-is a goal and it is unreasonable to assume that students will have
-already mastered syntax and enough understanding of the development
-environment (interpreting compiler messages, debugging, etc) then CLT
-would suggest that extensive /scaffolding/ be employed to reduce
-extraneous cognitive load. Being careful to sequence simple-to-complex
-concepts will help reduce /intrinsic/ cognitive load: i.e. teach bare
-objects and make sure students have began to automate tasks associated
-with implementing bare objects before introducing inheritance.
+% Keep in mind that providing scaffolding for syntax and grammar,
+% e.g. in the form of worked examples, or implementation in which
+% students need to copy/paste implementation code into object template
+% code may help with teaching OOP concepts, but would not be aiding
+% learning of actually implementing OOP code from scratch. Simply seeing
+% the implemented code is likely not enough to aid most students in
+% learning this skill themselves and if that is a goal of the course, or
+% a subsequent course it is important to focus on that while scaffolding
+% other aspects (e.g. providing an object interface and description of
+% how the object is intended to function, ask students to write
+% implementation). The main challenge with implementing such an
+% instructional strategy is that it will likely require coordination
+% across courses and instructors as teaching both programming, OOP,
+% debugging strategies all at once will likely not be met with much
+% success. A possible exception might be if one considers a spiral
+% curriculum, but the success of this still depends on close
+% coordination across courses and instructors.
 
-Keep in mind that providing scaffolding for syntax and grammar,
-e.g. in the form of worked examples, or implementation in which
-students need to copy/paste implementation code into object template
-code may help with teaching OOP concepts, but would not be aiding
-learning of actually implementing OOP code from scratch. Simply seeing
-the implemented code is likely not enough to aid most students in
-learning this skill themselves and if that is a goal of the course, or
-a subsequent course it is important to focus on that while scaffolding
-other aspects (e.g. providing an object interface and description of
-how the object is intended to function, ask students to write
-implementation). The main challenge with implementing such an
-instructional strategy is that it will likely require coordination
-across courses and instructors as teaching both programming, OOP,
-debugging strategies all at once will likely not be met with much
-success. A possible exception might be if one considers a spiral
-curriculum, but the success of this still depends on close
-coordination across courses and instructors.
-
-The authors of the Lohani et al report describe efforts to use the
-Alice programing environment to teach introductory programming
-concepts\autocite{lohani_reformulating_2011}. In theory a graphical
-programming environment should aid in the , learning of higher level
-concepts since it eliminates the extraneous cognitive load of worrying
-about syntax and grammar. In fact, there has been success in using
-iconic languages to teach first year programming
-\autocite{calloni_iconic_1997}, and research indicates that teaching
-graphical programming first does not inhibit a student from later
-learning textual programming languages\autocite{hundhausen_can_2009}.
-
-Eric Raymond describes three different kinds of complexity in software
-design: essential complexity, optional complexity, and accidental
-complexity
+% Eric Raymond describes three different kinds of complexity in software
+% design: essential complexity, optional complexity, and accidental
+% complexity. This has interesting parallels to the different types of
+% cognitive load in CLT. In particular the notion of \emph{essential
+%   complexity}, like \emph{intrinsic} cognitive load is associated with
+% the the solution itself: a solution to a complex problem will be
+% essentially complex due to the complexity associated with the problem
+% itself.
 

### 2016-05-02 paper.tex
about half way through with proof reading and minor edits

@@
 paradigm, learning to program may actually be a way to learn to
-problem solve, organize knowledge, and conceptualize processes.
+problem solve, organize knowledge, and reason about processes.
 
@@
 \autocite{engelbart_augmenting_1962}, a much more expansive view than
-simply becoming tools to automate tedious calculations. What was meant
-by this was that computing allows for new ways of dealing with
+simply becoming tools to automate tedious calculations. In other words, computing allows for new ways of dealing with
 complexity, and provides new metaphors for thinking about
@@
 graduate may be beside the point, there are lasting cognitive benefits
-learning to interact with a computer can provide.
+that learning to interact with a computer can provide.
 
@@
 indicate a call to action
-\autocite[e.g.][]{mccracken_multinational_2001,kolling_problem_1999},
+\autocite{mccracken_multinational_2001,kolling_problem_1999},
 or a survey of practices
-\autocite[e.g.][]{pears_survey_2007-1,robins_learning_2003} we used
+\autocite{pears_survey_2007-1,robins_learning_2003} we used
 the `cited by' list to find additional relevant papers.
@@
 Because the literature search of introductory programming resulted in
-variety of examples of how and why programming was taught, we needed a
+a variety of examples of how and why programming was taught, we needed a
 documented framework to consistently evaluate each one and guide our
@@
 (\citeyear{jamieson_creating_2009}) call to ground educational
-research activities in how people learn we identified the concept of
+research activities in how people learn, we identified the concept of
 \emph{constructive alignment} as a promising framework to evaluate and
@@
 
-\section{Current state of research}
+\section*{Current state of research}
 The bulk of literature about teaching introductory programming (CS1)
@@
 to support the learning objectives of the course. This behavior,
-making a popular practice part of a learning environment without
+adopting a popular practice into a learning environment without
 critically analyzing whether or not is a good fit, is by no means
@@
 confusion for students and frustration for
-instructor\autocite{kolling_problem_1999}.
+instructors\autocite{kolling_problem_1999}.
 
@@
 %% this could then be connected at the end to bullet points of suggestions
-\section{Trends in programming education research}
+\section*{Trends in programming education research}
 A \citeyear{robins_learning_2003} study by
@@
 students do not know how to program by the end of their introductory
-programming courses, sparked much conversation and push to reform
+programming courses, sparked much conversation and a push to reform
 introductory programming in the next decade. This phenomenon is by no
@@
 that despite the volume of work exploring the implementation of
-introductory programming courses ``there is little systematic evidence
+introductory programming courses, ``there is little systematic evidence
 to support any particular approach.''\autocite{pears_survey_2007-1}
@@
 
-\subsection{Components of a learning
+\subsection*{Components of a learning
 system}\label{sec:components-of-a-learning-system}
@@
 
-\begin{figure}[h!]
+\begin{figure}[!ht]
   \centering
@@
 
-\subsection{Effective learning systems}\label{sec:effective-learning-systems}
+\subsection*{Effective learning systems}\label{sec:effective-learning-systems}
 
@@
 
-\section{A hypothetical case study}
+\section*{A hypothetical case study}
 As noted, teaching OOP is a common area of interest in the
@@
 
-\subsection{Cognitive load theory}
+\subsection*{Cognitive load theory}
 Cognitive load theory (CLT) is based on the premise that the capacity
@@
 
-\subsection{Object oriented programming}
+\subsection*{Object oriented programming}
 %% Reviewer: more explicit connections to cogsci
@@
 
-\subsection{Cognitive challenges teaching OOP}
+\subsection*{Cognitive challenges teaching OOP}
 A behavior that characterizes expert programmers is the ability to
@@
 
-\subsection{Language choice}
+\subsection*{Language choice}
 Choice of language does matter: If the learning objectives of an
@@
 
-\subsubsection{Iconic versus textual languages}
+\subsubsection*{Iconic versus textual languages}
 
@@
 
-\subsection{Problem solving}
+\subsection*{Problem solving}
 Developing problem solving was widely regarded as a common objective
@@
 
-\section{Assessment}
+\section*{Assessment}
 %% Reviewer: short
@@
 
-\section{Considerations and limitations}
+\section*{Considerations and limitations}
 In the above hypothetical example we had the luxury of complete
@@
 
-\section{Discussion and future work}
+\section*{Discussion and future work}
 
@@
 
-\section{Conclusion}
+\section*{Conclusion}
 

### 2016-05-03 paper.tex
no need for relative path prefix
final proofread and edits

@@
 \usepackage[backend=biber,autolang=hyphen, bibencoding=inputenc, isbn=false, uniquename=false, sorting=none, sortcites=true, style=numeric, autocite=superscript, url=false]{biblatex} % biblatex setup
-\addbibresource{./computing.bib}
+\addbibresource{computing.bib}
 \appto{\bibfont}{\footnotesize} % for 10pt bibliography
@@
 that despite the volume of work exploring the implementation of
-introductory programming courses, ``there is little systematic evidence
-to support any particular approach.''\autocite{pears_survey_2007-1}
-We think this could be due to two reasons:
+introductory programming courses, ``there is little systematic
+evidence to support any particular
+approach.''\autocite{pears_survey_2007-1} We think this could be due
+to two reasons:
 
@@
 
-\item Even if goals were clearly stated it is likely that many
+\item Even if goals were clearly stated, it is likely that many
   different approaches could be used successfully if chosen
@@
 used in procedural programming\autocite{kolling_problem_1999}. As
-should not be surprising there are some problems for which OOP
+should not be surprising, there are some problems for which OOP
 provides a good set of tools to solve in an efficient, elegant manner,
@@
 depends on the specific learning goals. The role of the instructional
-designer then is to reduce extraneous load and intentionally managing
-germane load.
+designer then is to be aware of intrinsic load, reduce extraneous load
+and intentionally managing germane load.
 
@@
 systems\autocite{kay_history_1996}. The underlying metaphore of Kay's
-conception for OOP was the biological cell: each object is
-designed to accomplish a specific set of tasks where the complexity of
-these tasks is encapsulated within the object. Objects interact with
-one another by sending and receiving messages, and systems of these
+conception for OOP was the biological cell: each object is designed to
+accomplish a specific set of tasks such that the complexity of these
+tasks is encapsulated within the object. Objects interact with one
+another by sending and receiving messages, and systems of these
 objects are assembled to solve complex
@@
 themselves combined with the cognitive load of learning to author a
-program in a particular environment, do not exceed the learner's
+program in a particular environment, does not exceed the learner's
 working memory capacity.
@@
 In a superficial way, CLT theory would suggest that OOP is a helpful
-paradigm for managing cognitive omplexity: objects encapsulate
+paradigm for managing cognitive complexity: objects encapsulate
 complexity by hiding the details of implementation behind a simple
@@
 calls is more challenging than grouping lines of imperative codes into
-re-usable procedures. Thus to be fluent in OOP design a programmer
+re-usable procedures. Thus to be fluent in OOP design, a programmer
 \emph{must} be able to switch between thinking at the object layer and
@@
 with writing code and thinking about state and methods fills up a
-learners working memory they may have difficult learning concepts of
-abstraction, even when the instructor explicitly points them out. To
-reduce the cognitive load associated with writing code and working
-with object state and methods, these tasks must be practiced until
-somewhat automated: they can be done without conscious
-thought\autocite{ambrose_chapter_2010}.
+learner's working memory, the learner may have difficulty grasping
+concepts of abstraction, even when the instructor explicitly points
+them out. To reduce the cognitive load associated with writing code
+and working with object state and methods, these tasks must be
+practiced until somewhat automated: to a point at which they can be
+done without conscious thought\autocite{ambrose_chapter_2010}.
 
@@
 complex syntax features\autocite{koulouri_teaching_2014}. The
-explanation for this can be found in CLT. If, for example, the goal is
-to teach problem-solving techniques, there will be cognitive load
-associated with thinking about different strategies one could
+explanation for this can be found in CLT. Given the current example of
+a goal to teach problem-solving techniques, there will be cognitive
+load associated with thinking about different strategies one could
 use. Learning the syntax rules of a programming language adds to the
@@
 
-In theory an iconic programming environment such as
+In theory, an iconic programming environment such as
 Scratch\autocite{resnick_scratch_2009} or
@@
 environment, focus group data indicated a dissatisfaction with Alice
-as an introductory language resulting in a switch to
-LabVIEW\autocite{lohani_reformulating_2011}.
+as an introductory language, resulting in a switch to
+LabVIEW\autocite{lohani_reformulating_2011}. A possible explanation
+given for the dissatisfaction of Alice was that students did not
+perceive it as being a ``real'' programming environment they might use
+in industry, while LabVIEW was. Certainly, helping students connect
+the utility of skills they learn in the classroom to those that will
+be important in their later career is important. In some cases
+switching to a tool that is known to be used in industry may be
+useful, though this approach seems shortsighted: it is unlikely that
+all graduates will end up in an industry that uses a particular tool
+(e.g. LabVIEW), so if that is the motivation for a particular tool we
+must still address the utility of this tool to students who may never
+have reason to use it again. A different strategy than swathing tools
+is to better articulate \emph{why} a particular tool was chosen for
+the learning environment. Students might have less resistance to an
+educational environment such as Alice if they know that it can help
+them learn high level concepts that \emph{will} be directly applicable
+to their future career \emph{and} that the use of Alice as an
+introductory environment will not hinder them from later transferring
+those concepts to other environments.
 
 \subsection*{Problem solving}
-Developing problem solving was widely regarded as a common objective
+Developing problem solving skills was widely regarded as a common objective
 for introductory programming courses. Achieving this objective is more
@@
 when they begin programming. Second, it helps manage cognitive
-load. Where as in the previous example of language choice we reduced
+load. Where as in the previous example of language choice, we reduced
 cognitive load by choosing a language with a simpler syntax, here we
@@
 system. Designing good assessment is
-hard\autocite{lister_blooming_2000,traynor_synthesis_2005}, and often good assessment can
-require significant time on the part of the instructor. Because of
-this, assessment is often the weak point, lowering the effectiveness
-of the entire system.
+hard\autocite{lister_blooming_2000,traynor_synthesis_2005}, and often
+good assessment can require significant time on the part of the
+instructor. Because of this, assessment is often simplified, lowering
+the effectiveness of the entire system.
 
 If our course objective is to teach learners to solve problems with
-programming then an ideal assessment would be to give students a
+programming, then an ideal assessment would be to give students a
 problem to solve, ask them to write a program to solve it as well as
@@
 courses\autocite{lister_blooming_2000,lister_objectives_2001,sheard_exploring_2011},
-and in particular automated
+and in particular, automated
 assessment\autocite{cheang_automated_2003,traynor_synthesis_2005,ihantola_review_2010}. In
@@
 constructive alignment framework would constrain the types of course
-objectives and instruction that were feasible. For instance, if the
-primary course objective was to teach conceptual understanding of
+objectives and instruction that would be feasible. For instance, if
+the primary course objective was to teach conceptual understanding of
 computer science, then a well-designed multiple-choice assessment
 could be used
-effectively\autocite{lister_blooming_2000,lister_objectives_2001}.
+effectively\autocite{lister_blooming_2000,lister_objectives_2001}. Using
+the framework of constructive alignment to critically evaluate the
+alignment of assessment, objectives, and instruction, may help make
+explicit some of the compromises that must be made for larger class
+sizes. This positions course designers to be more intentional about
+what compromises to learning are made.
 
@@
 demands may shift focus to languages such as Erlang that are designed
-for high levels of concurrency as well as demand we adopt different
-metaphors for computing\autocite{stein_challenging_1999}.
+for high levels of concurrency, result in more interest in functional
+paradigms over OOP, as well as demand we adopt different metaphors for
+computing altogether\autocite{stein_challenging_1999}.
 
@@
 
-Recognizing that introductory programming is a common component of
+Recognizing that introductory programming is a common component of an
 engineering curriculum, and that there is a national push to introduce
@@
 technologies they learn are no longer relevant. We strongly advocate
-for the use of a framework such as constructive alignment, described in
-this paper, to inform the choices for course objectives, instructional
-techniques, and assessments. Using such a framework well help identify
-components of a course that may be contributing unnecessarily to
-cognitive load, or reasons why student outcomes might not match
-expectations assumed from course objectives.
+for the use of a framework such as constructive alignment, described
+in this paper, to inform the choices for course objectives,
+instructional techniques, and assessments. Using such a framework well
+help identify components of a course that may be contributing
+unnecessarily to cognitive load, or reasons why student outcomes might
+not match expectations assumed from course objectives.
 
 Analyzing a proposed curriculum through the lens of cognitive load
-theory can aid in instruction decisions that will increase the chances
-achieving learning goals. Educators must critically review the goals
-of their introductory programming courses: if goals are to understand
-high level concepts such as OOP, but not necessarily implement these
-concepts in textual languages, it may be prudent to consider iconic
-languages for these courses. If however, goals include learning OOP
-concepts \emph{and} implementing these concepts in textual language,
-educators must be aware of the increased cognitive load associated
-with each of these tasks and provide scaffolding to help students
-automate aspects of each task individually before expecting them to
-combine them both in successful programs.
+theory can aid in instructional decisions that will increase the
+chance of achieving learning goals. Educators must critically review
+the goals of their introductory programming courses: if goals are to
+understand high level concepts such as OOP, but not necessarily
+implement these concepts in textual languages, it may be prudent to
+consider iconic languages for these courses. If however, goals include
+learning OOP concepts \emph{and} implementing these concepts in
+textual language, educators must be aware of the increased cognitive
+load associated with each of these tasks and provide scaffolding to
+help students automate aspects of each task individually before
+expecting them to combine both in successful programs.
 
 Taken together, using an established framework, such as constructive
-alignment, to design an analyze curriculum, and utilizing cognitive
+alignment, to design and analyze a curriculum, and utilizing cognitive
 load theory to help reduce extraneous cognitive load while being
-intentional about inclusion of germane cognitive load should help
+intentional about inclusion of germane cognitive load, should help
 instructional designers create learning environments in which students
@@
 have to be adjusted to account for the finite limitations of working
-memory capacity: it is likely not feasible to expect students with no
-programming experience to learn the syntax and grammar of a new
-language \emph{and} higher level concepts associated with a paradigm
-like OOP, \emph{and} be able to switch between abstraction layers in
-the fluid manner necessary to solving complex problems with
-programs. What this suggests is that closer coordination is needed
-across courses in a curriculum so that goals of one become background
-knowledge of subsequent courses, resulting in successful achievement
-of program learning goals by the time of graduation.
+memory capacity: it is likely not feasible in the span of a single
+semester to expect students with no programming experience to learn
+the syntax and grammar of a new language \emph{and} higher level
+concepts associated with a paradigm like OOP, \emph{and} be able to
+switch between abstraction layers in the fluid manner necessary to
+solving complex problems with programs. What this suggests is that
+closer coordination is needed across courses in a curriculum so that
+goals of one become background knowledge of subsequent courses,
+resulting in successful achievement of program learning goals by the
+time of graduation.
 
